---
title: "Childhood family structure in 16 European countries"
author: "Andreas Gustafsson"
institute: "Stockholm University"
format: 
  revealjs:
    embed-resources: true
slide-number: true
progress: true
touch: true
reference-location: margin
reference-links: TRUE
bibliography: bibliography.bib
css: "presentation_style.css"
---


```{r}
#| include: false
#| warning: false
#| echo: false

# include library and targets source
library(targets)
library(fst)
library(ggplot2)
library(forcats)
library(data.table)
library(kableExtra)
tar_config_set(store = "../_targets/")
source("../functions/figure_functions.R")
```

## Introduction

::: {.incremental}
* The late 20^th^ and early 21^st^ century brought about large changes in family dynamics
* From nuclear families to a diversity of family forms
* Large differences between European countries and by socioeconomic status
:::

## Research questions

1. What are the most common family structures among European children, and how do they vary across European countries and regions?
2. How does the risk of experiencing a particular family structure differ by the mother's socioeconomic status, as measured by level of education?

## Previous research {.smaller}

::: {.incremental}
* European differences [@andersson_life-table_2017]:
    * Intact families more common in Southern and South-Eastern Europe
    * Single-mother births are the most common in Central-Eastern Europe
    * Experiencing parental separation is the least common in Southern and South-Eastern Europe. But also among certain Western European countries (Germany & Netherlands)
* Parental education is associated with certain processes [@kalmijn_new_2021], but unclear connection to others [@turunen_entering_2011;@dahlberg_pathways_2018;@jalovaara_disparities_2018]
:::


## Data

:::: {.columns}

::: {.column width="40%"}
* Harmonized Histories
* Child histories from maternal partnership and childbearing history
:::

::: {.column width="60%"}
```{r}
#| echo: false

format_table(
    tar_read(children_filtered),
    format = "html",
    response_rates = file.path("..","data", "response_rates.csv")
) |>   kable_styling(font_size = 12)

```
:::

::::

## Methods

::: {.incremental}
1. Create maternal partnership trajectory from birth until age 15
2. Calculate sequence dissimilarities
3. Cluster data
4. Cluster diagnostics
5. Show cluster distributions
:::


## Final clusters

```{r}
#| echo: false
tar_read(p_index)
```

## Regional proportions

```{r}
#| echo: false
tar_read(p_region_p)
```

## Educational associations

```{r}
#| echo: false
tar_read(p_educational_p) + theme(axis.title.y = element_blank(), aspect.ratio = 0.35)
```

## Conclusions

* South/North divide
* Stratification by maternal education
* Important to follow children across multiple transitions

## References