---
format: 
  pdf:
    documentclass: article
    papersize: a4
    geometry: "left=2.5cm, right=4cm, bottom=2cm, top=2cm"
    mainfont: Times New Roman
    sansfont: Verdana
    fontsize: 12pt
    subparagraph: yes
    include-in-header:
      text: |
        \usepackage{booktabs} 
        \usepackage{xcolor} 
        \usepackage{siunitx}
        \usepackage{lscape}
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{geometry}
        \usepackage{fontspec}
        \usepackage{ragged2e}
        \usepackage[british]{babel}
        \usepackage{setspace}
        \usepackage{etoolbox}
        \AtBeginEnvironment{table}{\sffamily}
        \usepackage{fontspec}
        \usepackage{titling}
        \usepackage{sectsty}
        \sectionfont{\huge\fontspec{Verdana}}
        \subsectionfont{\Large\fontspec{Verdana}}
        \subsubsectionfont{\normalsize\fontspec{Verdana}}
        \usepackage[labelfont=bf,sf]{caption}
        \usepackage{pdfpages}
    include-before-body:
      text: |
        \includepdf{front_page.pdf}
        \onehalfspacing
        \RaggedRight
        \hypersetup{linkcolor=black}
        \tableofcontents
        \listoftables
        \listoffigures
        \hypersetup{linkcolor=NavyBlue}
        \newpage
    latex-engine: xelatex
link-citations: true
urlcolor: blue
reference-location: block
csl: elsevier-harvard.csl
bibliography: "bibliography.bib"
---

```{r}
#| include: false
#| warning: false
#| echo: false

# include library and targets source
library(targets)
library(tarchetypes)
library(fst)
library(ggplot2)
library(forcats)
library(data.table)
library(kableExtra)
tar_config_set(store = "../_targets/")
knitr::opts_chunk$set(
  out.width="100%",
  dev="png",
  fig.dpi = 600
)

knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",")
})

tar_load(children_clusters)

```


# Introduction

The European family dynamics of the late 20th and early 21st century has undergone large changes as compared to previous decades. In particular through increasing family complexity. From the amassed family demography literature, the majority of attention has been placed on investigating the family complexity generating processes of divorce, re-partnering and multipartner fertility from the adult perspective and less is known about the family transitions of children. Having a good picture of to what degree and when children experience different family circumstances are important as both family structure and transitions have the potential to influence child wellbeing and future outcomes [@fomby_family_2007]. There is a vast literature on both proximate and long term effects of family structure on a diverse set of outcomes. For instance, recent finding indicate that proximate effects of living in a stepfamily can range from lower test scores in primary school [@lee_family_2022] to insomnia [@nilsen_sleep_2022] to excess externalizing problems [@mostafa_impact_2018]. At the same time these effects are contested and an earlier review has found mixed support for the association between family structure and child outcomes [@hadfield_changes_2018]. 

The demographic studies that have been carried on children's experiences of family transitions in Europe [such as @andersson_childrens_2002;@kennedy_childrens_2010; @thomson_family_2014], have primarily looked at the degree to which children experience particular events. Albeit important, transitions are only one part of the family experience. Increasing attention has been placed on the importance of also studying duration as well as the sequence of transitions. This has been done withing the US literature on childhood family experiences [@mitchell_pathways_2013;@johnston_two_2020] using profile based methods such as latent class analysis and sequence analysis but so far it has not been done on a representative sample of the European child population. In order to fully understand the changes in family structure changes during childhood we are in need of methodological advancements to better capture potential diversity.

Further, there has been ample research on how these family demographic changes have developed in Europe, and how the processes have differed between European regions. This has often been approached through the Second Demographic Transition theory. Despite this rich flora of comparative research, very little has been done to document the implications for children's family trajectories and to what degree they differ between countries. Open questions that remain to be answered are to what degree children's family experience mirror adult experiences and if the same processes that drive changes in adult family trajectories also affect children in the same fashion. In terms of divorce, it seems that the increasing rate of separations among adult couples have not been mirrored among children, and that it is highly dependent on parental educational level [@kalmijn_new_2021]. In some European countries, children of highly educated parents have not experienced an increased risk of parental separation at all [@kalmijn_new_2021]. 

The objective of this thesis is to provide a descriptive analysis of children's family structure trajectories in Europe. In particular I will answer the questions (1) What are the most common family trajectories among European children, and how do they vary across European countries and regions? (2) How does the risk of experiencing a particular family trajectory differ by the mother's socioeconomic status, as measured by level of education? This will be accomplished by generating a set of family history sequences for European children, based on the Harmonized History data set. Analysis will be performed by sequence analysis, through which I will generate clusters of the most common family trajectories of European children. I will then perform a descriptive analysis of the proportion of children who experience these trajectories on a country by country basis. Further, I will describe how the most common trajectories are associated with the mother's socioeconomic status. This will contribute to our current knowledge on children's family lives by describing how their family structure changes over time, relating it to both country of residence and socioeconomic background. Thus we will gain further knowledge on how children's family lives vary both within and between countries.

Before describing the methods in greater detail, I will outline the main theoretical frameworks of this thesis as well as previous research on children's experiences of family structure changes. In the end I will present the results of my analysis and put them into context in the discussion and conclusion sections.

# Theory and previous research 

## Family demographic drivers of childhood family structure

The three childhood states that I will be analyzing in this study, _living with original parents_, _living with a single parent_ and _living in a stepfamily_, arise due to circumstances and decisions made by the child's parents. So in order to understand the family conditions of children, we first need to know which processes at the parental level  shape family forms. How common these processes are among adults (and among parents in particular) will determine how common different family structures are for children. In this section I will thus outline the parental processes that lay behind why children are in one state or another as well as provide summary measures of how common they are in Europe and their relationship to socioeconomic status. In the next section I will provide more detail on how the intensity of these processes differ between European regions. 

### Living with original parents

The large majority of European children are born into a union where the adults are also the biological parents of the child. and only a small percentage of these have another adult outside the union as their biological parent. 

For the first state, _living with original parents_, it can arise from a set of different processes. Commonly, we think about being born into a parental union where both parents are the biological birth parents. There are however other processes that give rise to children being born into a union. Assisted reproductive technologies, in particular egg and semen donation, can lead to a child being born into a union where one of the parents is not the biological parents. Paternal discrepancy is also not uncommon, where the presumed biological father of a child is not the actual biological father. Apart from discussing the mechanisms here, I will not make any difference between the different routes behind being born into a union within this study. 

Population wide data of maternal union status at child birth in Europe between 1975 and 2004 indicate that between 80-97% of births are within a union, with large differences between countries [@perelli-harris_changes_2012]. However, good quality population estimates of paternal discrepancy and the number of children conceived through sperm donation, egg donation or surrogacy are lacking. In a review of previous studies (n = 17) @bellis_measuring_2005 found that rates of paternal discrepancy in Western countries varies between 0.8 and 30%, with the median being 3.7%. Regarding the use of assisted reproductive technologies, 1.9 to 5.9% of European infants in 2010 were conceived through ART, depending on country [@kupka_assisted_2014]. Around 0-20% of these births arouse through egg donation and 13% (total European average) from sperm donors (own analysis of supplementary material), indicating that total births where one birth parent is not a biological parent is still rare through ART. In sum, the large majority of European children are born into a union and only a small percentage of these have another adult outside the union as their biological parent. 

### Living with a single parent

Living with a single parent is a state that can arise from three main processes: birth to a single mother, parental separation and the death of a parent. Births to single fathers through surrogacy exists, but is so far very limited [@jones_single_2023]. In terms of single mother births, we have already indicated that single mother births account for 3-20% of all births in Europe during the time period 1975 to 2004 [@perelli-harris_changes_2012]. One important predictor that has been found cross-nationally in Europe is that there is a negative educational gradient associated with single mother births [@perelli-harris_educational_2010]. Lower education is associated with an increased risk in giving birth as a single mother compared to giving birth in a union. This gradient can to some degree be explained both by a lower likelihood of entering a partnership as pregnant as well as less access to contraceptives and abortions [@koops_socio-economic_2021].

Children being born into a union may later experience living with a single parent due to parental separation. This may be the result of both separation from a cohabiting union or divorce from a marriage. In terms of divorce, it has been steadily increasing in Europe across a wide set of countries during the latter half of the 20th century [@kalmijn_explaining_2007;@kalmijn_new_2021;@harkonen_divorce_2014]. Estimates of net divorce rates (annual divorces per 1000 married women) in Europe between 1990 and 2000 ranged from 6.41 to 20.35 [@kalmijn_explaining_2007]. A review by @amato_divorce_2010 found that important predictors of divorce were female employment and female income. This is hypothesized to stem from that these resources make women less dependent on men for their sustenance, thus opening up the possibility for divorce. The same review also found that the association of education and divorce risk vary across countries, and later studies have also found that the country context seem to matter more than education for explaining the prevalence of union dissolution [@perelli-harris_partnership_2016]. Union dissolution may also occur when a non-marital cohabiting union ends and dissolution rates of cohabitation has been found to be consistently higher than for marriages [@andersson_life-table_2017]. However, the consequences for children does not directly reflect those of adults in this sense, as the presence of children within the union at the time of dissolution is much lower for cohabiting unions than for marriages [@galezewska_cross-national_2017].

The third process which may lead to living with a single parent is the case when the other parent passes away. Single country studies have found that the percentage of children that experience the loss of a parent are around 3% in Sweden [@hiyoshi_childhood_2020] and 5% in the UK [@akerman_childhood_2011] but there are no cross European population measures on the prevalence of childhood bereavement. 

Finally, it is important to note that separation does not always entail that children end up with just one single parent. Joint physical custody has become more common, were children at times live both with their mother and their father after separation. In most European countries, around 0-10% of children live in a symmetric joint physical custody arrangement when the family is not intact [@steinbach_joint_2021]. Sweden is an outlier where the occurrence is 20%. In the case were a parent is the sole carer of a child, this is most often the mother. However it has been estimated, using that European Community Household Panel Survey, that in 1996 around 10% of European lone-parents were fathers [@chambaz_lone-parent_2001].

### Living in a stepfamily

For children to experience living in a stepfamily, it requires that the single parent forms a new cohabiting partnership or marriage. This may occur either through re-partnering after separation or when forming a partnership after being a lone birth parent. There are no cross European studies on the partnering dynamics of lone birth mothers with them as the unit of analysis, but taking the child perspective @andersson_life-table_2017 show the percentage of children born to lone mothers who are in a stepfamily before the age of 6. There the percentages range from 25% to 55% depending on European country, thus it is a quite common event but by age 6 the majority of children born to lone mothers remain with a single mother. The partnering opportunities of lone mothers are not necessarily equal withing countries. @kalmijn_poverty_2010 finds that among lone birth mothers in the Netherlands, the risk of entering a union is higher for mothers with average income than for those who are poor. Contrarily, @jalovaara_disparities_2018 finds that the reverse patterns is true for Finnish children, where children to poor mothers are more likely to experience living in a two-parent family. Thus, the link between socioeconomic status and union formation for lone mothers is not necessarily the same in European countries.

As for entering a new union after union dissolution, @galezewska_cross-national_2017 finds that in 2005 around 60-80% of European women repartner within ten years since first union dissolution. The estimates are also similar either if the first union is a marriage or cohabiting union. As for the social determinant of re-partnering, findings from Estonia, Lithuania, Poland and Hungary indicate that education does not seem to be associated with women's risk of repartnerning, but that education can play a role for men in the propensity to repartner [@maslauskaite_gender_2015]. However, @dahlberg_pathways_2018 finds that for countries in Northern and Western Europe, mothers with post-secondary education are more likely to repartner than mothers with secondary education and lower. This pattern is also found among fathers but is generally weaker. So as with union formation after being a single parent, education is not a consistent predictor of the risk of entering a new partnership. 

## Mapping European differences

As we saw in the section above, the prevalence of family states and the rates of transitions differ between European countries and in this section I will outline these in more detail. In order to describe these patterns I will divide the Europe into five geographic regions which partly have a shared history as well as common family demographic trends. These regions are are Northern, Western, Southern, South-Eastern and Central-Eastern Europe. This scheme has been used in the same, or a very similar format, in many previous studies on family dynamics [see for example @billari_leaving_2001;@fokkema_trends_2008;@kalmijn_explaining_2007;@thomson_family_2014;]. In the western parts of Europe, there has been a consistent North-South divide in family tie patterns [@reher_family_1998] which can be expected to influence family transitions. There is however a point in separating out the Nordic countries as they have been seen as forerunners in terms of family demographic changes [@ohlsson-wijk_family_2020]. In terms of the Eastern European countries there is much diversity in family behaviors and they have been divided in up to 5 distinct regions in previous research [@thornton_sweeping_2009], but in this paper I will adopt a two-part division into South-Eastern and Central-Eastern [as in @kalmijn_explaining_2007]. 

### Northern Europe

### Western Europe

### Southern Europe

### South-Eastern Europe

### Central-Eastern Europe

## Previous studies on children's family structure

The variety of childhood family structures and transition have been documented to some extent in Europe from a child-centered perspective. A general patterns that has been found is that children from socioeconomically disadvantaged groups experience family disruptions to a larger degree. In Sweden, children are more likely to experience parental separation if their parents are less-educated [@kennedy_childrens_2010]. This is also echoed in @kalmijn_new_2021, where similar results were found also for Belgium, Czech Republic, Estonia, France, Hungary, Norway and the United Kingdom. They also found the this differential risk of parental divorce has been increasing since the 1980's. When it comes to stepfamily formation, @turunen_entering_2011 found no effect of parental education on the risk of entering a stepfamily after having lived with a single parent (either due to separation or being borne to a lone parent) for Swedish children. Thus, the risk of entering a complex family seems to be largely dependent on the different separation risks that children experience.

It is important to note that there are large cross-national variations in risks of parental divorce and stepfamily formation in Europe. Using life-table methods, @andersson_life-table_2017 describe these risks for a large set of European countries. From their descriptive analysis we can see that the risk of parental separation by age 15 is highest in parts of the post-Socialist countries (Russia 42%, Estonia 36%, Lithuania 35% but also France 35%) while its the lowest in Southern Europe and the southern parts of post-Socialist Europe (Georgia 10%, Bulgaria 12%, Spain 12% and Italy 12%). There are no clear patterns when it comes to the proportions experiencing a stepfamily after parental separation but it varies from 14% (Georgia and Italy) up to 60% (Belgium). As for births to single mothers, who are neither in cohabitation or marriage, The Czech Republic and Russia stands out with 13% of children experiencing this. But there are also high levels in France, Estonia, Lithuania and Poland (8% of all children in these countries.) On the other side of the spectrum we have Sweden (2%), Norway (3%) and Netherlands (3%).

Other aspects which are important to consider when tracing children's family experiences, both in the US and in Europe, is if cohabitation is included as a measure as well as the prevalence of joint physical custody and father-only households. Relying on the US National Survey of Family Growth,  @rally_cohabitation_2004 found that when adding family transitions into and out of cohabitation, family instability (calculated from transitions into and out of parental partnerships) increased by 30% for White children and over 100% for Black children. This highlights the importance of including cohabitation in studies on family complexity. For all countries studied, @andersson_life-table_2017 also found that experiencing a parental separation is more common for children born into cohabiting unions. As for joint physical custody, in a majority of European countries it is a fringe arrangement that only 5% or less of children experience, whilst it is quite common in Sweden (21%) and Belgium (13.5%) [@steinbach_joint_2021]. This makes it hard to examine the actual family experiences of children based on surveys where parents are the anchors or where children's experiences are not tracked after separation on both the maternal and paternal side. If the prevalence was similar across countries it would be simpler to approximate children's experiences based on parental reports. Current praxis of following children's family trajectories on the maternal side will now under report the total experienced complexity more in countries such as Sweden and Belgium whereas it will be a good approximation in countries where joint physical custody is uncommon. For instance, estimates of the proportion of children with half-siblings have been found to be lower if there is fertility information only on one parent [@cancian_evolution_2011]. There are findings from the UK which show that children's family complexity (measured with a compound complexity index) are similar between father's and mother's family lives over childhood [@rowold_cohort_2022], indicating that a fair share of complexity is missed when only accounting for children's experiences on the maternal side. Attempts to ameliorate this issue have been made [see @kalmijn_multi-actor_2018] but is so far uncommon. 

In a review of the research on family diversity and complexity among adolescents, [@pearce_increasing_2018] argues that to better understand both the family circumstances of adolescents as well as how family characteristics are related to certain outcomes, we are in need of new methods. They suggest, among others, the use of "profile based" methods, such as latent class analysis and cluster analysis. So far, very few studies have been conducted on children's family structure using this type of methodology but we will look more closely at two studies that have. 

Using the U.S. National Longitudinal Survey of Youth, @mitchell_pathways_2013 created a five type classification of children's family structure on the maternal side using latent class analysis. The model distinguished between living with (1) married biological parents, (2) cohabiting biological parents,  (3) cohabiting stepfather, (4) married stepfather and (5) single mother. The five class solution created typologies that Mitchell named, "Married biological parents who divorce", "Long-term single mothers", "Continuously married biological parents", "Cohabiting biological parents who marry or break up" and "Gain a stepfather". This study can be credited as an early attempt at finding childhood family trajectories but are in some senses lacking. One of the largest drawback is making a distinction between cohabiting parents and married parents. Although this might be important from an adult perspective, it probably makes no difference in how children perceive the parental relationship. It further gives rise to the category "Cohabiting biological parents who marry or break up", which really makes no substantive sense. Classifying children who experience a separation into the same group as those who might experience a strengthening of the parental relationship seems to be just an artifact of how the model and it's states are constructed. Therefore, we need to be aware of how model construction can influence the outcomes with this type of data-centric analysis. 

A more recent study, using the children of the participants in the National Longitudinal Survey of Youth, performed and compared classifications of family trajectories through both latent class analysis and sequence analysis [@johnston_two_2020]. This study did not distinguish between cohabitation and marriage and revealed some further patterns in family structure changes, and where each pattern was intuitively interpretable. In particular, the latent class analysis distinguished classes on the basis of timing, where early and late entry into both a family with a sole mother as well as into a stepfamily became distinct types. It was also found that the different types identified both by sequence analysis and latent class analysis predicted the risk of later adolescent alcohol consumption. 

## The consequences of family structure

Complexity in the family domain has been defined as the combination of family structure and sibling composition during childhood [@manning_family_2014]. Where a "simple" family may be defined as a household with both parents present and a few number of full-siblings, a complex family may consists of step-parents as well as half- and step-siblings. The four main demographic drivers of complexity are birth to lone mothers, separation and divorce, re-partnering, and multi-partner fertility [@thomson_family_2014]. Phenomena which have all become more salient during the latter 20^th^ century. Understanding how family structures change during childhood is important because family transitions and different family arrangements have tangible consequences for the children involved. In this section I will outline what some of these consequences are, focusing both on the consequences of experiencing particular events as well as living in certain family arrangements.

One of the most well studied childhood family transition is divorce, along with its consequences on proximate and future child outcomes. These consequences occur in multiple domains, ranging from well-being to educational outcomes. @averdijk_parental_2012 found that parental separation lead to an increase in both aggressive and internalizing behavior among children in Zurich, controlling for both parental conflict, maternal depression and financial difficulties. The effect was partly mediated by maternal depression. Similar results were found in the U.S context, where children of divorce, who were propensity-score-matched to children of intact families, showed higher levels of both internalizing and externalizing behavior following parental separation [@weaver_mediation_2015]. This effect was found both in the short- and long-term following separation. There are also potential health consequences of divorce, where German children experiencing divorce, studied using a panel setup, were found to increase their BMI after divorce [@brockmann_unhealthy_2013]

There are also consequences on school outcomes. @havermans_family_2014 found an association between divorce and less school engagement among Belgian children. Large part of the effect was mediated by the parent-child relationship, financial problems at home and parental conflict. Swedish children experiencing divorce have also been found to show decreased psychological well-being and lower educational attainments [@gahler_parental_2015]

In opposition to divorce, the event of re-partnering has also been studied, but not to the same degree. @bachman_is_2009 studied adolescent from low-income urban US families, and found that neither risks nor benefits increased as a result of maternal marriage on a wide battery of adolescent well-being indicators. Although introducing a new partner in the household increased financial resources, it was not associated with improvements in home-environment or maternal well-being, which could explain why adolescent functioning did not increase significantly. In a further US study, children's academic achievement was not found to be improved when their single mother married [@wagmiller_does_2010]. The small benefits to academic performance was also to a large degree distributed among the children to mother's with higher education. There is yet little evidence for the case that maternal re-marriages increases the life chances and well-being of their children [@hadfield_changes_2018].

Apart from the outcomes of family transitions, there is also a literature on the effects of the long term states of living with a single mother or in a stepfamily, contrasted with living in a nuclear family with original parents. For the case of single mother families in the U.S, @fomby_family_2017 found that children born to lone mothers showed higher levels of behavioral problems than those born into nuclear families. Youth in single mother families in Germany have also been found to experience lower well-being than those in living with their mother and father, or their mother and a stepfather [@walper_effects_2015]. There are also external factors to consider, which might influence the well-being of children in single mother families, one example comes from Sweden where living in a single parent household has been found to be associated with a higher risk of bullying for children [@laftman_national_2017].

All in all, these findings suggest that both family transitions and long-term family states matter for the well-being of children in several dimensions. A consequence of this is also that studies that focus on either transitions or states will miss out the other process, indicating a need for a approaches taking both of these factors into account in the study of children's family lives. This has also been highlighted by @pearce_increasing_2018, who argue that to go further in our understanding of the associations between family complexity and its related outcomes, we are in need of new methods. A set of method which they suggest as a potential way forward are "profile-based" methods, such as latent class analysis and sequence analysis. 
<!-- 
## Life course theory 

According to Elder et al. [-@elder_emergence_2003], life course theory came about from a "desire to understand social pathways, their developmental effect, and their relation to personal and social-historical conditions". Since its incipience in the 1970's, the *life course* has grown from a scientific idea to a theoretical framework and even a research field in its own right. With the launch of *Advances in Life Course Research* as a quarterly journal, it was declared that the life course is coming of age [@billari_life_2009]. 

The main theoretical concepts from the mature life course research which will be used in this thesis is those of *transitions*, *trajectories*,  *pathways* and *linked lives*. *Transitions* are state changes that are discrete and time-bound [@george_sociological_1993]. These can occur in multiple domains --- such as the social, biological, or economic --- and examples that have already been discussed are divorce and re-partnering. In turn, multiple transitions make up life *trajectories*. These consists of longer periods of change and stability, and although marked by a series of transitions they have a continuity of direction [@hutchison_life_2005]. A family formation trajectory by Hollywood blueprint would consist of dating followed by cohabitation and finally a wedding, marking a transition into married life. 

Just as trajectories are made up of a set of transitions, *pathways* are constituted by interlocking transitions and trajectories across time and domains [@macmillan_families_2005]. In particular, pathways are often in relation to entering and exiting social institutions such as a career or family. Although the combinatorial possibilities of transitions and trajectories are vast, the number of actualized pathways are usually limited and in relation to normative systems [@white_application_2014]. A way of thinking about pathways is that they are lived in relation to schema; that is, blueprints of culturally appropriate behavior, which are enacted more or less consciously [@sewell_theory_1992]. Schema inform individual heuristics for the temporal organization of the life course by providing a set of rules and models for how to link pathways over time [@macmillan_families_2005]. A modern Scandinavian pathway in relation to family formation would be forming a cohabiting union followed by having children, skipping the previously normative transition of marriage [@hofacker_patterns_2014]. 

Although often emphasized in the literature, the power of normative schema in shaping the individual life course has been contested [@macmillan_structure_2005]. Apart from norms, institutional and structural processes also shape lives and are able to produce homogeneity in life patterns [@marini_age_1984], as we have also seen in the discussion on the proliferation of the SDT. A further way of thinking about pathways, uncommitted to theoretical assumptions about their generation, is simply as a set of probabilities of transitioning between different states. But rather than a set of independent probabilities, they are path dependent with a certain degree of historicity, which can give rise to common patterns in a population [@rutter_pathways_1989]. 

Finally, a central idea in the life course framework is that of *linked lives*. This concept highlights the interdependence of our lives and that transitions that occur in one persons life often entail transitions for other people [@elder_emergence_2003]. This principle has particularly been applied in studies on the family and intergenerational relations [@carvalho_family_2021]. Already in one of the first studies using a life course perspective, @elder_children_1974 found that children's development could be impeded by the economic hardships experienced by their parents. This interdependence between parents and children, and how their social lives are intertwined, has since been explored in much depth and breadth. 

It is with this interdependence in mind, that I approach my object of study. Children's early family experiences are highly dependent on their parents' partnership trajectories, both in terms of vertical and horizontal family relations. At the same time, children's early family trajectories may themselves be associated with future outcomes. Thus, it is of interest to know how children's family trajectories are constituted. Before turning to how to conceptualize and operationalize these trajectories, I will turn to a conceptual model that can be used to understand why childhood family life course trajectories may differ between European countries. 

## Explaining cross European differences

One of the most adopted theories in explaining the macro level changes in family patterns during the late 20^th^ century is the Second Demographic Transition (SDT) theory. The SDT theory starts by identifying a set of phenomena, such as rising levels of cohabitation, living apart together arrangements, higher divorce rates, fertility postponement and an increase in out-of-wedlock childbirths [@lesthaeghe_second_2014]. In the early SDT theory, the changing family dynamics were seen as a result of ideational changes in Western Europe. In particular, the rise of postmodern attitudes and norms [@van_de_kaa_postmodern_2001], tightly linked to Inglehart's concept of a post-materialist society [@inglehart_silent_1971]. In particular, ideas such as individualism and self-realization are seen to have lead to both more individualized decisions on family formation, uncoupled from previous family norms, as well as a greater acceptance for issues such as homosexuality, abortion and non-marital childbearing. 

Ideational changes do not however, exist in a vacuum. In a recognition that also other social forces shape family decisions, Lestaeghe and Vanderhoeft proposed the conceptual model of _Ready, Willing, and Able_ [@lesthaeghe_ready_2001], adopted directly from @coale_demographic_1973. _Readiness_ refers to that new forms of behavior must come with some utility to the actor and _willingness_ to the acceptability of new patterns of action in relation to normative schemes, which is the aspect usually associated with the SDT. _Ability_ on the other hand, refers to if condition are in place to realize these new forms of behavior. In the case of family life, ability is related to, for instance, access to contraception and legal regulations of family relationships. This conceptual model bridges the relation between ideational changes, established normative systems as well as technological and legal systems. As both norms and institutional factors differ between countries, we would expect that even if changes in individual attitudes uniformly move towards more post-materialist ideals, the degree to which these attitudes can be realized will differ between countries and the processes have, as we shall see, been very different. 

Scandinavia has been a forerunner with regards to changing family forms. In Sweden, childbearing began to precede marriage already in the in the mid 1970's, when childbearing within cohabiting unions started to get mainstream acceptance [@thomson_partnership_2005]. At the same time, there was a drastic decrease in the level of total first marriage rates, dropping from about 0.9 in 1965 to 0.6 in 1975 [@sobotka_changing_2008]. In 2008, 60% of all children where born in non-marital cohabiting relationships [@olah_sweden_2008]. Drivers have been hypothesized to include both individualism and residential autonomy as well as institutional support also for childbearing outside of marriage [@olah_sweden_2008].

In other parts of Western Europe the patterns have been similar. The decrease in total first marriage rate started later than in Scandinavia, but fell to an even lower level in 2005 [@sobotka_changing_2008]. In the Netherlands in 2000, 55% of the Dutch population would opt for cohabitation before marriage as their ideal relationship trajectory [@fokkema_netherlands_2008], indicating an ideational shift towards cohabitation as preferable over direct marriage. Similarly, marriage has been decreasing in France and in 2005, the number of births in non-marital cohabitation where almost the same as the number of births in marital unions [@toulemon_france_2008]. Similar patterns are also found in Austria where 2% of women born 1941-45 had their first child in a cohabiting union, compared with 18% for women born 1966-67 [@prskawetz_austria_2008]. Further, over 25% of Austrian women born 1971-81 consider marriage to be an outdated institution [@prskawetz_austria_2008].

In Southern Europe, the drop in marriage rates started in the late 1970s and has not been as pronounced as in Northern and Western Europe [@sobotka_changing_2008]. Further, non-marital cohabitation is not nearly as common as in Northern and Western Europe [@de_rose_italy_2008;@delgado_spain_2008]. Does this mean that the cultural shift in values and norms did not occur in these countries? According to previous findings this does not seem to be the case for either Italy [@de_rose_italy_2008] or Spain [@delgado_spain_2008]. Rather, it seems that the economic situation of youth, coupled with a lack of active family policies is inhibiting changes in family dynamics. Thus, the young opt for staying in the parental home for longer as the option of cohabitation with a partner is not economically feasible, which in turn has led to a postponement in union formation [@baizan_political_2002]. The Southern European case seems to be a good example of where individuals display a high level of _willingness_, but lack both the resources (_readiness_) and proper support (_ability_) to change their family behaviors; which in turn gives rise to a different demographic pattern than in Northern and Western Europe.

Another region where there are clear mismatches between readiness, willingness and ability is in the Post-socialist countries. We first see evidence of the SDT in this region after the fall of the Iron Curtain in the early 1990s, when marriage rates started to drop steeply [@sobotka_changing_2008]. There is however a different timing when examining divorce, as already in the 1950's and 1960's parts of the Soviet block registered the highest divorce rates in Europe [@sobotka_changing_2008], indicating that divorce was an acceptable phenomena already in the early phase of the SDT in parts of Eastern Europe. It has also been found that tolerance towards pre-marital sex and divorce as well as values such as self-fulfillment outside of the family became salient in countries such as Poland [@kotowska_poland_2008] and Lithuania [@stankuniene_lithuania_2008] in the early 2000s. 

Tracing the developments in this region through the Readiness-Willingness-Ability approach, @sobotka_overview_2008-1 finds mixed indications in all three dimensions. The development of a social security net and the widespread employment of women increased readiness already before the fall of the Iron Curtain, but at the same time these jobs were often low-payed and females were still expected to take care of the household. As for willingness, the anti-religiousness of the Communist ideology helped break down traditional family norms and induced an atomization of society through the breakdown of religious and civic organizations [@speder_hungary_2008]. However, these new values were often adopted selectively [@sobotka_overview_2008-1]. Legal systems often enabled some of the family changes of the SDT, but contraceptive technology was often lacking showing that the ability to adopt new family behaviors was only possible to certain degrees.

Summing up the different regional patterns described above, two main pathways of the SDT have been identified and described [@sobotka_overview_2008-1]. The first path was driven by the classic SDT narrative of individualization and self-fulfillment, and primarily occurred among the affluent and highly educated. The second path arose as a response to changed structural conditions and economic crises, and primarily occurred among the less advantaged groups. Here there is also a regional divide, where the first pathway was primarily taken by countries in Northern, Western and to some degree Southern Europe, whilst Post-socialist countries mainly took the second.  -->

## Operationalizing family structure over the life course

As has been outlined, the family structure of children may exhibit many different patterns and if we want to fully understand this complexity, we need to study how these patterns play out over time. One fruitful analytic method which has been used within the life course literature is sequence analysis [for an overview see @liao_sequence_2022]. The foundation of sequence analysis lays in describing temporal processes as a series of states and state changes over discrete time units (such as months or years). It thereby provides a format for describing individual life course trajectories.

We can expect that the structures that children experience are a finite yet potentially complex mix of the three family states _living with original parents, _living with a single parent_ and _living in a stepfamily_. We can expect two main different structural pathways, the first is for children born into a union and the second for children born to two parents. Following this initial family constellation children born into a union may experience parental separation and eventually also stepfamily formation. Children born to a lone parent can experience stepfamily formation without any intermediate family state. Following the first stepfamily, children may also experience multiple parental dissolutions and new stepfamilies. Since this can yield a multitude of family structures over time, we need a way to reduce this complexity to a comprehensible form. One way of capturing how these structural changes and finding the most common patterns are through cluster analysis. 

Cluster analysis is a data-centric approach which deviates from traditional statistical methodologies in two main ways: 1) the units of analyses and 2) data assumptions. @aisenbrey_new_2010 show these differences by contrasting event history analysis with sequence analysis. Where event history analysis studies probabilities of transitions, sequence analysis studies patterns in trajectories as their main units of analysis. When it comes to assumption about the data, event history analysis assumes that the data is generated by a stochastic process whereas the algorithmic clustering makes no assumptions about the underlying process. This also leads to differences in practice, where event history analysis usually employs models where included covariates are derived from theory or previous empirical research, sequence analysis relies on algorithmic clustering of the data to identify groups of interest. These clusters are then used in two main ways: most commonly they are used as a dependent categorical variable in multinomial regression to find how other covariates predict life trajectories [@piccarreta_holistic_2019] but the clusters can also be used as predictor variables to see how life trajectories influence later life outcomes [e.g. @johnston_two_2020].

However, a fair amount of critique has been leveled against the clustering techniques employed by sequence analysis. To put this critique in methodological context, I will first outline the general steps in producing sequence analytic clusters before discussing the methodological drawbacks. The first step is to create a distance (or dissimilarity) matrix by calculating the distance between each pairwise combination of individual sequences in the data. Since the seminal work of @abbott_optimal_1986, the Optimal Matching algorithm has provided a standard way of calculating such a distance metric. The basic intuition is that we can permute a sequence by two kinds of operations to arrive at another sequence. The operations are substitutions (replacing one state by another) and insertions/deletions of states (commonly known as _indels_). In most cases, it is however not as straightforward as just summing up the number of operations needed to permute one sequence into another to get their distance. As the probability of transitions between states differ between pairs of states, or between substitution and indel operations, we need a substitution matrix which gives a cost to each possible substitution between pairs of states. We also need to assign a cost to indel operations. Thus, the optimal matching algorithm tries to find a set of operations that optimize the cost of permuting one sequence to another. This optimized cost is used as a measure of distance. Using the finalized distance matrix, it is possible to apply a clustering algorithm to derive a set of clusters, this is commonly done through Ward's method [@aisenbrey_new_2010]. This is an agglomerative hierarchical clustering algorithm that produces cluster by minimizing the within-cluster variance, using the squared dissimilarity distances as inputs [@murtagh_wards_2014].

The criticism leveled towards this methods is focused on two main areas: firstly, on the perceived lack of substantial results produced by the methodology and secondly, and perhaps more importantly, on the relationship between methodological assumptions and social processes. Both @wu_comments_2000 and @levine_but_2000 criticize sequence analysis for not delivering on early promises and that substantial results and new areas of research have not been borne out of it. As summarized in @liao_sequence_2022, there is a large variety of sociological articles being published using sequence analysis which would render that point of contention moot, but it does emphasis the importance of scrutinizing the sequence analysis methodology and if it is a valid tool for modeling or describing social processes.

@wu_comments_2000 offers a critique of sequence analysis where the most pressing is if a discrete sequence with events coupled with substitution and indel costs is a good model of life trajectories. The main problem lies in determining costs, as each cost is an assumption the researcher has to make, and that costs in the optimal matching algorithm are symmetric. This symmetric assumption means that transition from state $s_j \rightarrow s_k$ carries the same cost as transition from $s_k \rightarrow s_j$. But the question posed by Wu, is if the transitions between social states, such as employment and unemployment, are symmetrical in their probabilities or not. Which would imply that sequence representation through optimal matching is not isomorphic to the social processes they purport to represent. Further, Wu argues that as determining substitution and indel costs is up to the discretion of the researcher, it introduces more assumptions and potential biases in the analysis than proponents of sequence analysis had up to then acknowledged. This is a valid criticism but loses some of its relevance if we consider sequence analysis as an exploratory or descriptive analysis method, rather than a substitute to statistical modeling.

## Aims and expectations

The aim of this thesis is to further our understanding of children's family lives in Europe by examining them through the entire childhood life course. By using the profile based methods of sequence and cluster anlyses I aim to distill the diversity of family trajectories to a set of distinct family trajectory types. By doing this I will be able to compare the children of different countries and create a descriptive picture of childhood family lives in Europe and to further understand their similarities and differences. Given that within country differences are also to be expected, I will aim to see to what extend the mother's socioeconomic status is related to the family trajectory of her children. In particular I will try to answer the research questions:

  1. What are the most common family structures among European children, and how do they vary across regions and countries? 
  2. How does the risk of experiencing a particular family structure differ by the mother's socioeconomic status, as measured by level of education?

Given the descriptive character of this paper I will not do any formal hypothesis testing but in light of the theories and previous research outlined above, I have some expectations of what will be seen in the data. As for the first research question, I expect that we will see the smallest variation in trajectory cluster membership in Southern Europe and the southern countries in post-Socialist Europe where a large majority of children will live their entire childhood in intact families. In the norhtern Post-Socialist countries, I expect there to be a larger share of non-intact family trajectories which will consist of both children born to lone mothers and to children of divorce. In Scandinavia I expect that trajectories driven by lone mothers will be rare, but that there will be a sizable share of children of divorce. Similar patterns are expected for Western Europe, perhaps with a smaller degree of stepfamilies and somewhat larger degree of single-mother families compared to Scandinavia. As for the second research question, I expect that the proportion of children with intact family histories will increase with increasing education. In particular, I expect that the incidence of non-intact families driven by birth to single mothers will decrease with education. 

# Data and Methods

## Analytical strategy

The first step in the analysis consists of defining family states and generating individual family sequences. As the data does not track both original parents of a child, I will only follow the mother's partnership trajectory with the assumption that children are residing at least part time with their mother during their childhood. This is an assumption I have to make as the data does not contain information if the child resides with the responding parent. Thus, joint physical custody and father only households are not taken into account, and the family structure on the father's side are not measured. This will probably make the measures of family complexity biased downwards. A positive note is that it will avoid the problem that men tend to under report their number of children in retrospective surveys [@juby_where_1999]. So although an analysis solely on the mother's side is a limitation, we can proceed as long as we keep that limitation in mind. Having constructed the parental structure sequence, I compute sequence dissimilarities and generate clusters for the full study population. The final number of clusters are selected using both cluster quality criteria and the substantive significance of the cluster division.

The problems of the validity of clustering typologies have been recognized for a long time, and methods for ameliorating this deficit have been developed within the bioinformatics literature [for an overview see @von_luxburg_clustering_2010], few sequence analysis studies within the social sciences have taken this issue to heart. In order to improve the validity of the typology, I will use a parametric bootstrap technique proposed by @hennig_flexible_2015 and implemented for sequence data by @studer_validating_2021 to aid with the selection of the final clusters.

Once the family typology is in place, I will show the proportion of the different family types for each country and region in the study, thus providing a descriptive of how children's family trajectories vary across Europe. This will also be broken down further by the mother's educational level.

## Data

The data used in this study comes from the birth and union histories in the Harmonized Histories data set [@perelli-harris_harmonized_2010]. This data set is compiled from a set of national surveys, primarily the Generations and Gender Survey, but also with other nationally representative surveys on partnerships and childbearing. The data is based on individual retrospective reports on partnership and childbearing histories. The information on children is based on biological children so I do not analyze adoptees in this thesis. I also use information on highest achieved education, represented as a 3-level scale which has been constructed from ISCED 1997. The levels are primary (ISCED 0-2), secondary (ISCED 3-4) and tertiary educaiton (ISCED 5-6).

The surveys were conducted differently in different countries. Generally, they were conducted through interviews [@fokkema_generations_2016], but self-filled web questionnaires were also used [@instituto_nacional_de_estadistica_fertility_2019]. Different sampling procedures were also used and more detailed information can be found in @fokkema_generations_2016 and @perelli-harris_harmonized_2010. Response rates are avaialble for most surveys and are presented below. 

To construct my analytical sample from the Harmonized History dataset, I start of by excluding countries from the data set. Germany is excluded due to its poor family history data and the UK is excluded due to oversampling respondents with low socioeconomic status. Non-European countries are also removed. I then have `r tar_read(parents_tagged)[,.N]` total observations from `r length(tar_read(parents_tagged)[,unique(COUNTRY)])` different countries. I remove all men from the sample. Next, I remove childless women and respondents with missing birth information. I remove those where there is missing information on time of union or dissolution, overlapping relationships and those where a union dissolution is reported to have occurred before the union start. 

In order to be able to analyze the trajectories from the children's lived experience I reshape the data so that each child becomes a unit of observation. At the start I have `r tar_read(children_tagged)[,.N]` children. I remove children with missing birth month or where the birth is reported to have occurred before the mother is born. As I want to track children's' experiences until the end of age 15, I remove those in the sample who have not turned 16 at the time of interview as well as those who left home or died before their 16^th^ birthday. As I am interested in family changes after the advent of the Second Demographic Transition, I only keep those children born after 1980. In total `r tar_read(final_data)[,.N]` children, to `r tar_read(final_data)[unique(RESPID),.N]` mothers, remained in the sample. 

In @tbl-1 I show basic information about the study population. For all countries in the study children born between 1980 and 1988 are included. For some countries such as Belgium, Netherlands, Poland, Spain and Sweden there are also children born well into the 1990's and the early 2000's. Each country, except Austria, is well represented with over 1,000 children. Belarus, Norway, Poland and Spain all have over 3,300 children included. That over 12,000 of the `r tar_read(final_data)[,.N]` respondents come from the Catholic countries Poland and Spain, might bias the analysis towards favouring clustering solutions which works well for these particular family contexts. Which is a discussion I will return to later. Apart from this, there is a well balanced sample with good representation from all European regions and enough respondents to perform subgroup analyses. 

```{r table1}
#| echo: false
#| results: "asis"
#| label: tbl-1
#| tbl-cap: "Descriptive information on survey years, response rates, cohorts included and number of children from each country. Some response rates are missing due to the sampling procedure of the surveys."

tar_read(table_1)
```

## Sequence generation

For each child a family history was constructed on the maternal side, which is composed of the mother's partner history during the child's first 15 years of life. In order to construct this history, a STate-Sequence format [STS, @gabadinho_analyzing_2011] is used. This means that for each time point of the child's life a column is generated which is filled with values describing the mother's union status. The process time for analysis is from birth until turning 16, with quarterly precision. In STS format this is represented by 60 consecutive states, each capturing a three month period. 

Firstly, union spells were constructed. As there is no information on who is the father of the child if a mother was partnered at time of birth, that partner was assumed to be the other main parent. Then for each quarter, the representative state was coded as either *Single*, *Original parents* or *Step* depending on the mother's partner status in the first month of each quarter. The state is coded as *Single* if the mother is not in a union during the first month of the quarter. It is coded as *Original parents* if the mother is still in a relationship with the same partner as at the time of birth. Finally, a state is coded as *Step* if the mother is with a different partner than at time of birth. All consecutive partners are assumed to be new, i.e. no re-partnering with a former partner is assumed to occur. I do not distinguish between opposite-sex and same-sex partnerships, and the original parents are not necessarily the same as the biological parents, but simply constitute the partnership that the child is born into. Formally, the maternal union sequence can be described as the vector $\textbf{S}$ where  

$$
\textbf{S} = (s_1,s_2,s_3, \cdots, s_{60}) | s_i \in \{\textrm{Original parents, Single, Step} \}
$${#eq-sequence-example}

An example sequence of a child's first two years of life, where the mother is first partnered for 12 months (four quarters), then single for 6 (two quarters) and finally repartnered for 6 months (two quarters) is visible in @tbl-state-ex.

```{r}
#| echo: false
#| results: "asis"
#| label: tbl-state-ex
#| tbl-cap: "Two year union state sequence example"

months <- paste0("", 1:8)
states <- c(
  "OP", "OP", "OP", "OP",
  "Si", "Si", "St", "St"
)
knitr::kable(
  rbind(
    months,
    states
  ),
  "simple",
  row.names = FALSE
)
```

## Analysis

Having constructed the maternal union sequence, I calculate a dissimilarity matrix using an Dynamic Hamming Distance algorithm. This algorithm requires that the sequences are of the same lengths and does not use insertion and delation operation. A benefit of this approach is that it increases the sensitivity to sequence timing [@lesnard_setting_2010], which we know is important from the life course literature. It also means that indel costs needn't be defined which reduces the amount of decisions, and potentially biases, when generating the dissimilarity matrix. The substitution costs are derived from the transition rates in the data and thus reflects transition probabilities in this particular sample. This is the most data-centric approach possible for defining substitution costs and also reduces potential researcher biases, with the drawback of potentially reducing the external validity of the final clustering solution. 

This dissimilarity matrix is used to cluster the sequences, which is done using Ward's method. Being an agglomerative method, it starts with $k$ number of clusters where each observation in the sample is its own cluster, thus $k = N$ in the first step. The algorithm then progressively merges clusters until there is only one cluster containing all $N$ observations. Larger clusters are formed in steps by merging the two clusters with the shortest distance, where distance is defined by the increase in sum-of-squares if the clusters are merged. Given that this clustering method produces solution between $1$ and $N$ clusters, a choice regarding the final number of clusters needs to be made. This choice will be based on sociological relevance and cluster quality indices.

In choosing the number of clusters, it is of interest to maximize internal cluster homogeneity as well as between-cluster differences, and a common measure of this propriety is the Average Silhouette Width , introduced by @kaufman_finding_1990. This measure produces a value between $-1$ and $1$, where a larger value indicates a stronger structure. I also use Hubert's C , which captures the discrepancy between the partition obtained and the theoretically best partition, given the particular number of groups and their distances [@hubert_general_1976]. Hubert's C is defined in the range $[0,1]$ and a smaller value indicates a better partitioning. The issue with assessing cluster quality is that, unlike in traditional statistical modelling, there is no null model to compare the results with. So, even though @kaufman_finding_1990 propose an interpretation of the Average Silhouette Width measure, it cannot be generalized to all cluster applications. 

A potential solution is to use parametric bootstrapping to generate _null-model like_ distributions, cluster them, calculate a cluster quality index and compare it to the quality index value of the clusters from the actual data [@hennig_flexible_2015]. This has recently been applied to sequence analysis [@studer_validating_2021] and to my knowledge, this study will be one of the first to apply this technique. In practice, I generate a data set of 50 sequence sets that are similar but non-clustered using a combination of both sequencing and timing parameters [for the detail of the process see @studer_validating_2021]. Then the Average Silhouette Width and Hubert's C are calculated for each cluster solution ranging from 2 to 10 clusters for each of the 50 bootstraps. This produces a confidence interval for each of the quality measures for each $k$ grouping. This provides a baseline measure of the quality indices for data that is not clustered. If the index values of the cluster solutions for the actual data falls out of this bootstrapped range, it is a strong indicator that the particular clustering solution represents a "true" cluster and not just an artifact of the clustering process. However, that a cluster is true from a purely data-centric view, does not mean that it holds any form of social significance. Therefore, apart from using cluster quality indices, I will also look at the theoretical basis of the best performing cluster solutions when selecting the number of clusters to use. Further, I only consider 2 to 10 clusters as a larger number of cluster would be both unwieldy in terms of discussing the number of groups and the number of members in the smaller clusters would be too few to carry out any form of description on the country level. 

I provide a visual summary of the sequence clusters by a set of three plots for each cluster. First, a state distribution plot which shows the relative frequency of each state in the population by month. Second, a modal plot which shows the relative frequency of the most common state for each time point. Third, a medoid plot which shows the sequence with the smallest average dissimilarity to all other sequences in the group. It can be seen as a "median" sequence, or as the most typical trajectory within the cluster group. I also label the clusters with qualitative descriptions that aim to capture the general family experience of each cluster. 

In the next step, I provide a description of the proportions of children which belong to each cluster, on a country level. This allows us to detect similarities and differences between European countries as it relates to children's family experiences. In the final step I also relate parental union histories to the mother's educational level to see if there are also patterns stratified by socioeconomic status and if these are similar or different between countries.

The analysis is performed in R and relies on the packages **TraMineR** and  **WeightedCluster** [@studer_weightedcluster_2021] for the sequence and cluster analysis. 

## Ethical considerations

The Harmonized Histories data set is de-identified to protect the personal integrity of the study participants and all analysis presented are as aggregated results. In order to get access to the data I have signed a pledge of confidentiality, pledging not the share the data and only perform analysis related to the approved project. The data collection has been approved in each individual country and has been done under the oversight of the GGP Ethics Board and its Data Protection Office. I do not deal with any directly sensitive personal data such as ethnicity, political opinions or union membership. Thus, this paper has small to no likelihood of harming survey participants in any way. On the other hand, we can not expect there to be any direct benefits to the survey participants either. The indirect benefits, which are shared with the rest of the European population, may come from an increased understanding of childrens family structures and how it relates to both country of residence and socioeconomic status. In the long term, this research may contribute a small piece of understanding on what drives childrens' family conditions, which is a first step to improving them.

While I do not analyze any sensitive personal data, I think it is important to acknowledge that my research subject may be sensitive in certain contexts. Blended families, stepfamilies and single mothers are still stigmatized across Europe and the question of family relations, both in terms of norms, values and legal protection is a debated political topic. As such, I avoid making normative claims surrounding particular family configurations and try to remain objective and close to the data when describing family trajectories, especially in light of mothers' socioeconomic backgrounds. 

In order the improve both the transparancy and reproducibility of my research all code for generating the results, as well as the thesis, is available as a replication package at GitHub [@gustafsson_addegezonfamily_complexity_in_childhood_2022]. Anyone wishing the reproduce this research will have to apply for the Harmononized History dataset and then it will be possible to re-run my exact analyses using the replication package. I have used version control for both the code and the writing of the thesis, allowing any one interested to scrutinize the research process in great detail and with full transparancy.

# Results

## Cluster diagnostics

The results from the cluster diagnostic bootstrap is presented in @fig-cwi. The pink shaded area represents the $95%$ confidence intervals of the standardized Average Sillhoute Width and Hubert's C and the purple line indicates the cluster quality of the 2 to 10 cluster solutions using the actual data. The graph shows that all cluster solutions in the range $[2,10]$ lay outside of the bootstrapped span, which in turn indicates that, from a data-centric view, there is always a significant clustering structure.

```{r}
#| echo: false
#| label: fig-cwi
#| fig-cap: "Standardized Average Silhouette Width and Hubert's C for actual and bootstrapped data"

tar_read(p_cluster_cqi)
```

We also see that the top scoring solution for the Average Sillhoute Width is the 6-cluster solution, and Hubert's C the solutions are progressively getting better for each number of cluster above 5. Given that the Average Sillhoute Width suggests the 6-cluster solution and that this gives a solution with a fair amount of children in each cluster, I choose to use the 6-cluster solution in my analysis. Working with six clusters still allows us to uncover heterogeneity in the complex family trajectories while still getting enough power to perform subgroup analyses.


## Final clusters

The 6-cluster solution is visualized in @fig-index-plot. This is a sequence index plot that shows the unique sequences in each cluster. Each horizontal line represents a unique family trajectory as it changes over time. The labels of the clusters are my own qualitative classifications of the information displayed in the index plots. The number of sequences is the number of unique sequences in the cluster and the _weighted n_ is the total number of children belonging to that cluster.

```{r}
#| echo: false
#| label: fig-index-plot
#| fig-cap: "Sequence index plot of the 6-cluster solution. The horizontal axis represents time, spanning the entire childhood, and each line is an individual famiy trajectory. Each show different distributions of the states OP (Original Parents), Single (living with a single mother) and Step (living with mother and stepfather)."
#| fig-height: 8

tar_read(p_index) + 
  theme(
    axis.text = element_text(size = 7),
    strip.text = element_text(size = 7),
    plot.margin = margin(0,3,0,0)
  )
```

A first thing to note is that the "Intact original family" cluster is by far the largest one, with around `r round(children_clusters[cluster == "Intact original family",.N]/children_clusters[,.N] * 100, 0)`% of children in the full sample belonging to this cluster. On the opposite end of the spectrum we have the "Later stepfamily" cluster, which is composed of  `r round(children_clusters[cluster == "Later stepfamily",.N], 0)` children. Before showing the distribution of these clusters by country, I would like to spend some time to discuss each cluster and what information we can derive from these visualizations.

Starting of with the "Intact original family" cluster, we see that not all individuals actually live in an intact original family throughout childhood. A fair proportion (`r children_clusters[cluster == "Intact original family", round(sum(FAMILY_STATE177 != "OP")/.N, 2)]`) of children actually experience parental separation at some point, but all of them experience it after the age of 6. This highlights the need to be careful when infering group characteristics from sequence clusters.

The cluster "Original family to stepfamily" is composed of children who are all born into a parental union and spend at least some time in it before their parents divorce and eventually ending up in a stepfamily. In the index plot we can see that most children spend quite a short time with a single mother before she repartners, which is also highlighted in the medoid plot. We also see that a few children end up living with a single mother again towards the end of the observed period. 

In the third plot, "Separation", we observe a group of children where all are born into a parental union and the bulk experience a parental separation between the ages 6 and 9. A few experience stepfamily formation but most remain with a single mother throughout childhood. Here we can note that there is no "late separation" cluster, these few have instead been classified with the "Intact original family" group. The cluster "Single mother" consists mostly of children who spend their entire childhood with a single mother, but there are some who also live in a stepfamily or with their original parents in early childhood for a brief period of time. 

The fifth category, "Early stepfamily", seems to contain some heterogeneity within the cluster where we see two common forms of family trajectories. The dominant one, which I have named the cluster after, consists of children being born to a single mother and who then enter a stepfamily before the age of 10. There is also a small group who are born into a union but experience a divorce shortly afterwards. The sixth category "Later stepfamily" also contains these two types of trajectories, but here stepfamily formation occurs later in childhood. 

<!-- In @fig-complex-proportion we see the proportion of children that are in each cluster, on a country by country basis, grouped by region. The left part of the figure shows the proportion of all children and the right hand figure shows the distrubtion over only those children who are not classified as "Intact original family". This is so that we can see in better detail how the distribution looks of these family sequence types. Countries are sorted within their respective region by the proportion of children beloning to the "Intact original family" group.  -->

In @tbl-proportion-family-type we see the percentages of children belonging to each cluster by European country. Within square brackets I also show the 95% confidence intervals. A quick glance reveals that there are large variations between countries in how common membership in particular trajectory groups are, for instance membership in the "Intact original family" groups ranges from 72% of Austrian children to 90.5% of Italian families. 

```{r}
#| echo: false
#| results: "asis"
#| label: tbl-proportion-family-type
#| tbl-cap: "The proportion of children classified according to each family type by European country grouped by region, with 95% confidence intervals in square brackets."

tar_read(tab_proportion_family_type) |> 
  kableExtra::kable_styling(font_size = 7) |>
  kableExtra::column_spec(2:8, width = "1.7cm") 
```

What is also evident is that there are no obvious patterns related to the regions I defined. There seems to be much variation within the regions given. In order to discover potential geographical patterns, I calculate the relative risk of belonging the each cluster in each country, as compared to the European mean risk. This is shown in @fig-map-cluster. A value close to one indicates that the risk of belonging to a particular cluster is the same in that country as the average risk in Europe. A value higher than one (green colors in the figure) indicates that the cluster is more common in that country than the European average. So a value of 2 shows that the cluster is twice is common in a given country as compared to the European average. At the other end, values lower than one (purple colors) indicate that the cluster is less common than on average in Europe.


```{r}
#| echo: false
#| label: fig-map-cluster
#| fig-cap: "Relative proportion of children in a country belonging to a cluster (as compared to the European mean). Gray countries are not included in the analysis. "
#| fig-height: 10

tar_read(p_map_cluster) + 
  theme(
    text = element_text(size = 8),
    legend.title = element_blank()
  )  

```

Looking at these six maps, a pattern emerges that was not immediately visible in @tbl-proportion-family-type. There seems to be a general pattern that Italy, Bulgaria, Romania, Hungary and Poland always have proportions below the European mean for all clusters except for the intact family cluster. This appears to also be the case for Spain execpt for the clusters Early- and Late stepfamily. This gives us a South-South Eastern axis of similar cluster distributions. A second axis, although more diverse, is the Western to North Eastern axis comprised of the remaining countries with the exception of the Netherlands. In all of these, complex trajectories seem to be more common than in the South/South east but the proportion belonging to a particular trajectory type is not always similar. Two of the most polarizing clusters along these axis are _Original family to stepfamily_ and _Separation_, which are two clusters that are both driven by being born into a partnership and then experiencing a parental sepration. This appears to be more common along Western-North Eastern axis, and where the more Northern countries have an overrepresentation of children entering a stepfamily after sepration, whereas this seems to be less common in the other along this axis. 

In terms of the clusters that are driven by birth to single mothers (or very early separations), they are all less common along the southern axis with the exception of Spain, where children are more likely than the European average to experience stepfamily formation after being born to a single mother. Among the more central and northern countries, living with a single mother throughout childhood is more common than the European average with the exception of Austria, Sweden and Norway. But it is also common for children to experience stepfamily formation after being born to a single mother. The countries were a particularily high proportion of children experience any of the single mother-driven trajectories (_Single mother_, _Early_- and _Late stepfamily_) are Belarus, Lithuania and Austria with around 15% of children being classified in any of these trajectories.

As outlined in the theory section, we can expect these patterns to shift once we take the mother's socioeconomic status into account and therefore I also present the cluster distributions by maternal educational level. In @fig-map-edu this is visualized as a matrix of maps with the cluster on the y-axis and maternal educational level on the x-axis. The color of the countries is once again representing a kind of relative risk, but this time not between countries but rather between educational levels within countries. In this case, a relative risk of 1 indicates that the particular cluster is as common within that maternal educational level as it is in countries population as a whole, disregarding maternal education. A relative risk of 2 indicates that the particular cluster is twice as common among children to mothers of that particular educational level, in relation to the population as a whole. This illustrates if children of particular clusters are over- or underrepresented among mothers of low, medium or high educaiton for each individual country. The cluster  _Intact original family_ has its own scale for the plot colors in order for the differences between educational levels to be visible. The exact proportions can be found in @tbl-edu in the Appendix. 

```{r}
#| echo: false
#| label: fig-map-edu
#| fig-cap: "Relative risk of mother's having primary, secondary and tertiary education for each cluster as compared to educational distribution of the mother sample in each country. "
#| fig-height: 10
tar_load(p_map_edu_intact)
tar_load(p_map_edu_complex)

cowplot::plot_grid(
            p_map_edu_intact[[1]] + theme(legend.title=element_blank()),
            p_map_edu_complex[[1]] + theme(legend.title=element_blank()),
            ncol = 1, 
            rel_heights = c(1,3.3)
        )

```

Among those children in the _Intact original family_ cluster, we see a clear educational gradient for most countries. Children in this cluster tend to have tertiary educated mothers to a higher degree than average. One exception to this rule is Poland, where children in this cluster tend to have mother's with a secondary education to a higher extent than the sample average. Italy and Lithuania also follow a pattern where children in this cluster are overrepresented when their mother's have a primary or tertiary education, but underrepresented among mother's with secondary education. 

Moving to the _Original family to stepfamily_ cluster, we see that this cluster is overrepresented among children whose mothers have primary and secondary education among many of the Northern, Western and Central countries (Norway, Sweden, Estonia, Lithuania, Netherlands, Czech Republic, Austria and to some degree France). But for the Southern axis this cluster is underpresented among children to mothers with only primary education and is instead overrepresented among children who's mothers have secondary or tertiary education. For instance, the relative risk of children of this cluster to have a mother with secondary education in Italy is `r p_map_edu_complex[[2]][COUNTRY == "Italy" & cluster == "Original family to stepfamily" & EDU_3 == "Secondary", round(relative_risk,2)]`. The Polish case shows a similar pattern as in the case of the _Intact original family_-cluster. Also here, the cluster is overpresented among children to mothers with primary and tertiary education, but underepresented among children to mothers with secondary education.

For children in the _Separation_ cluster, they are generally underrepresented among mothers with only primary education, with the exception of Sweden, Hungary and Bulgaria. They are slightly overrepresented to mothers with secondary education in Spain, Italy and Norway and finally overrepresented when mother's have a tertiary education in Austria, Hungary, Romania, Spain, Sweden and Lithuania. For France, Poland and Estonia, there seem to be no to very modest effects of education. In general, there is no clear geographical or educational pattern that emereges for this cluster.

For the single mother driven clusters we can see that the cluster _Single mother_ is overrepresented among children to primary educated mothers in most countries, except for Spain, Italy, Austria and Romania. For Spain, Italy and Austria this overrepresentation is instead found amoung the secondary educated mothers. Among children to tertirary educated mothers, this cluster is overreprensted in Sweden, Poland, Czech Repulic, France, Italy and Bulgaria. Turning to the cluster _Early stepfamily_, we see the strongest universal educational gradient. This cluster is overrepresented among children to mothers with only primary education. This is generally the case also for the cluster _Later stepfamily_, with a few exceptions. Thus, there seems to be a strong educational gradient among the mother's to children who are classified as _Early_- or _Late stepfamily_, but it is not as clear among the children who live with a single mother that do not enter into a new partnership. 

In sum, we see that when comparing European countries without taking socioeconomic background into account there is a Northern/Southern divide in which family trajectories are common but once we take maternal education into account, much of this pattern disappers, and for most clusters (except for _Early_- and _Late stepfamily_) it is not replaced by an educational pattern but instead gives rise to a complex mosaic of cluster distributions.

# Discussion

\dots

## Limitations

There are a set of limitations with this study which are important to keep in mind when reading and interpreting results, as well as when designing further studies on children's family demography. The first limitation, which partly comes from the data and partly by design choice, is the removal of father's partnership histories as discussed in the method section. What is needed to ameliorate this is information on where the child resides after divorce. However, it would also require a more complex methodology as both parents partnership trajectories would have to be modelled in tandem for each child. 

Another set of limitations related to the data are the study period, overrepresentaiton of certain countries as well as response rates. The study period, in which I track children born between 1980 and 2003 with the bulk being born in the 1980's and early 1990's, can seem somewhat outdated and is perhaps a bit too early to really capture recent changes in family demographic behaviours. At the same time it captures an important time point when changes were on the way but perhaps had not yet fully manifested in all countries studied. In terms of overrepresentation of certain countries, this is primarily a potential issue for the cluster analysis where countries with more children represented will have a stronger weight in which clustering solutions are chosen. There is no clear way to fix this issue as using survey weights would have introduced the problem that larger countries would shape the clustering solution to a larger extent. Another cross country issue, is the one of survey response rates. Some response rates are relatively high around 70-85%, which is an indicator that the sample will be nationally representative. Other response rates, such as Italy at 19.1% or Poland at 33.%, are very low and thus the results from these countries need to be interpreted with caution in relation to their national validity. 

Finally, another limitation of this study is that I do not capture the entire family context of the children that I study. Apart from the parental structure, siblings are also an important component of family life. In particular in relation to changing parental dynamics, where we can expect changes in half- and stepsibling composition occuring during the time period that I study. I intentionelly left out siblings as this would have made the family trajectories way more complex, having to constuct one parental trajectory and one sibling trajectory. However, I do think this is a very important topic and suggestions for further studies would be to focus on a smaller set of countries (or only one) but instead taking the sibling history into account.

# Conclusion

\dots

# Acknowledgments

\dots

# References

::: {#refs}
:::

\setcounter{section}{0}
\renewcommand{\thesection}{\Alph{section}}

\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}

\setcounter{figure}{0}
\renewcommand{\thefigure}{A\arabic{figure}}

# Appendix {.appendix}

```{r}
#| echo: false
#| results: "asis"
#| label: tbl-edu
#| tbl-cap: "Percentage of children in each cluster, over country by maternal level of education."

tar_read(tab_education_proportions) |> 
  kableExtra::kable_styling(font_size = 7) |>
  kableExtra::column_spec(2:7, width = "1.7cm") 
```
