---
title: "Childhood Family Structure and Sibling Composition in 16 European Countries: A Sequence Analysis Approach to Family Complexity"
author: "Andreas Gustafsson"
date: \today
format: 
  pdf:
    documentclass: article
    papersize: a4
    geometry: "left=3cm, right=5cm, bottom=3cm"
    include-in-header:
      text: |
        \usepackage{booktabs} 
        \usepackage{xcolor} 
        \usepackage{siunitx}
        \usepackage{lscape}
        \usepackage{geometry}
        \usepackage{fontspec}
        \usepackage{ragged2e}
        \usepackage[british]{babel}
        \usepackage{setspace}
        \usepackage{titling}
        \pretitle{\begin{flushleft}\Large}
        \posttitle{\end{flushleft}}
        \preauthor{\begin{flushleft}}
        \postauthor{\end{flushleft}}
        \predate{\begin{flushleft}}
        \postdate{\end{flushleft}}
    include-before-body:
      text: |
        \onehalfspacing
        \RaggedRight
        \hypersetup{linkcolor=black}
        \tableofcontents
        \newpage
        \listoftables
        \listoffigures
        \hypersetup{linkcolor=DarkOrchid4}
    latex-engine: xelatex
link-citations: true
urlcolor: blue
reference-location: block
mainfont: "Happy Times at the IKOB New Game Plus Edition"
bibliography: "https://raw.githubusercontent.com/addegezon/bibliography/main/bibliography.bib"
---

```{r}
#| include: false
#| warning: false
#| echo: false

# include library and targets source
library(targets)
library(fst)
library(ggplot2)
library(forcats)
library(data.table)
library(kableExtra)
tar_config_set(store = "../_targets/")
knitr::opts_chunk$set(
  out.width="100%",
  dev="pdf"
)
```

# Introduction

The European family dynamics of the late 20th and early 21st century has undergone large changes as compared to previous decades. In particular through increasing family complexity. Divorce, remarriage and multipartner fertility have all increased since the "golden age" of marriage in the 1950's [@sobotka_changing_2008; @thomson_family_2014]. The result is a multitude of family forms and, in extension, uncertainties regarding the organization of family life. These processes has variously been labeled as the deinstitutionalization of marriage [@cherlin_deinstitutionalization_2004] or as the de-standardization of the life course [@bruckner_-standardization_2005]. One of the most successful explanatory frameworks describing these changes is the Second Demographic Transition Theory [@lesthaeghe_second_2014].

From the amassed family demography literature, the majority of attention has been placed on investigating the family complexity generating processes of divorce, repartnering and multipartner fertility from the adult perspective. Since it has been proposed that family transitions have a negative effect on the well-being of children [@fomby_family_2007] and because there is some evidence of this effect [@hadfield_changes_2018], it is important to understand to what extent children experience these and what the trends are. As have been shown, children's exposure to divorce events do not always mirror those of adults [@kalmijn_new_2021]. Demographic studies that have been conducted on children's experience of family disruption and parental repartnering [such as @andersson_childrens_2002;@kennedy_childrens_2010; @thomson_family_2014] have primarily looked at the degree to which children experience particular events. Albeit important, this does not fully capture the accumulated experience of family complexity over childhood. 

The objective of this thesis is to describe children's family trajectories and to classify these trajectories according to the most common types of childhood family experiences. This will be accomplished by generating a set of family history sequences for European children, based on the Harmonized History data set. Analysis will be performed by sequence analysis,  I will generate clusters of the most common family trajectories of European children. I will then perform a descriptive analysis of the proportion of children who experience these trajectories on a country level. In the final step I will describe how the most common trajectories are associated with the probability of having full- and half-siblings across different European regions. Before describing the methods in greater detail, I will outline the main theoretical frameworks of this thesis as well as previous research on children's experiences of family complexity. In the end I will present the results of my analysis and put them into context in the discussion and conclusion sections.


# Theory and previous research 

## Explaining cross European differences

One of the most adopted theories in explaining the macro level changes in family patterns during the late 20^th^ century is the Second Demographic Transition (SDT) theory. The SDT theory starts by identifying a set of phenomena, such as rising levels of cohabitation, living apart together arrangements, higher divorce rates, fertility postponement and an increase in out-of-wedlock childbirths [@lesthaeghe_second_2014]. In the early SDT theory, the changing family dynamics were seen as a result of ideational changes in Western Europe. In particular, the rise of postmodern attitudes and norms [@van_de_kaa_postmodern_2001], tightly linked to Inglehart's concept of a post-materialist society [@inglehart_silent_1971]. In particular, ideas such as individualism and self-realization are seen to have lead to both more individualized decisions on family formation, uncoupled from previous family norms, as well as a greater acceptance for issues such as homosexuality, abortion and non-marital childbearing. 

Ideational changes do not however, exist in a vacuum. In a recognition that also other social forces shape family decisions, Lestaeghe and Vanderhoeft proposed the conceptual model of _Ready, Willing, and Able_ [@lesthaeghe_ready_2001], adopted directly from @coale_demographic_1973. _Readiness_ refers to that new forms of behavior must come with some utility to the actor and _willingness_ to the acceptability of new patterns of action in relation to normative schemes, which is the aspect usually associated with the SDT. _Ability_ on the other hand, refers to if condition are in place to realize these new forms of behavior. In the case of family life, ability is related to, for instance, access to contraception and legal regulations of family relationships. This conceptual model bridges the relation between ideational changes, established normative systems as well as technological and legal systems. As both norms and institutional factors differ between countries, we would expect that even if changes in individual attitudes uniformly move towards more post-materialist ideals, the degree to which these attitudes can be realized will differ between countries and the processes have, as we shall see, been very different. 

Scandinavia has been a forerunner with regards to changing family forms. In Sweden, childbearing began to precede marriage already in the in the mid 1970's, when childbearing within cohabiting unions started to get mainstream acceptance [@thomson_partnership_2005]. At the same time, there was a drastic decrease in the level of total first marriage rates, dropping from about 0.9 in 1965 to 0.6 in 1975 [@sobotka_changing_2008]. In 2008, 60% of all children where born in non-marital cohabiting relationships [@olah_sweden_2008]. Drivers have been hypothesized to include both individualism and residential autonomy as well as institutional support also for childbearing outside of marriage [@olah_sweden_2008].

In Western Europe the patterns have been similar. The decrease in total first marriage rate started later than in Scandinavia, but fell to an even lower level in 2005 [@sobotka_changing_2008]. In the Netherlands in 2000, 55% of the Dutch population would opt for cohabitation before marriage as their ideal relationship trajectory [@fokkema_netherlands_2008], indicating an ideational shift towards cohabitation as preferable over direct marriage. Similarly, marriage has been decreasing in France and in 2005, the number of births in non-marital cohabitation where almost the same as the number of births in marital unions [@toulemon_france_2008]. Similar patterns are also found in Austria where 2% of women born 1941-45 had their first child in a cohabiting union, compared with 18% for women born 1966-67 [@prskawetz_austria_2008]. Further, over 25% of Austrian women born 1971-81 consider marriage to be an outdated institution [@prskawetz_austria_2008].

In Southern Europe, the drop in marriage rates started in the late 1970s and has not been as pronounced as in Northern and Western Europe [@sobotka_changing_2008]. Further, non-marital cohabitation is not nearly as common as in Northern and Western Europe [@de_rose_italy_2008;@delgado_spain_2008]. Does this mean that the cultural shift in values and norms did not occur in these countries? According to previous findings this does not seem to be the case for either Italy [@de_rose_italy_2008] or Spain [@delgado_spain_2008]. Rather, it seems that the economic situation of youth, coupled with a lack of active family policies is inhibiting changes in family dynamics. Thus, the young opt for staying in the parental home for longer as the option of cohabitation with a partner is not economically feasible, which in turn has led to a postponement in union formation [@baizan_political_2002]. The Southern European case seems to be a good example of where individuals display a high level of _willingness_, but lack both the resources (_readiness_) and proper support (_ability_) to change their family behaviors; which in turn gives rise to a different demographic pattern than in Northern and Western Europe.

Another region where there are clear mismatches between readiness, willingness and ability is in Eastern Europe. We first see evidence of the SDT in this region after the fall of the Iron Curtain in the early 1990s, when marriage rates started to drop steeply [@sobotka_changing_2008]. There is however a different timing when examining divorce, as already in the 1950's and 1960's parts of the Soviet block registered the highest divorce rates in Europe [@sobotka_changing_2008], indicating that divorce was an acceptable phenomena already in the early phase of the SDT in parts of Eastern Europe. It has also been found that tolerance towards pre-marital sex and divorce as well as values such as self-fulfillment outside of the family became salient in countries such as Poland [@kotowska_poland_2008] and Lithuania [@stankuniene_lithuania_2008] in the early 2000s. 

Tracing the developments in Eastern Europe through the Readiness-Willingness-Ability approach, @sobotka_overview_2008-1 finds mixed indications in all three dimensions. The development of a social security net and the widespread employment of women increased readiness already before the fall of the Iron Curtain, but at the same time these jobs were often low-payed and females were still expected to take care of the household. As for willingness, the anti-religiousness of the Communist ideology helped break down traditional family norms and induced an atomization of society through the breakdown of religious and civic organizations [@speder_hungary_2008]. However, these new values were often adopted selectively [@sobotka_overview_2008-1]. Legal systems often enabled some of the family changes of the SDT, but contraceptive technology was often lacking showing that the ability to adopt new family behaviors was only possible to certain degrees.

Summing up the different regional patterns described above, two main pathways of the SDT have been identified and described [@sobotka_overview_2008-1]. The first path was driven by the classic SDT narrative of individualization and self-fulfillment, and primarily occurred among the affluent and highly educated. The second path arose as a response to changed structural conditions and economic crises, and primarily occurred among the less advantaged groups. Here there is also a regional divide, where the first pathway was primarily taken by countries in Northern, Western and to some degree Southern Europe, whilst countries in Eastern and South-Eastern Europe mainly took the second.

Informed by this sections, I will group the countries in my data into Nordic, Central, Eastern and Southern Europe for parts of the analysis. These are regions which have seen similar trends as captured by the SDT. Using this grouping will give more numerical power to the analysis as well a better pedagogical summary of the diverse results produced. However, we need to keep in mind that this will hide some of the diversity between countries. 

Up until now, I have mainly focused on the relationship between the SDT and the life course of adults, but it is not hard to envision that as adults family lives change, so do the family lives of children. Thus, we turn to the discourse on family complexity to understand how children have been affected by the processes previously outlined.

## Family complexity: predictors and consequences

<!-- complexity in "family complexity" sense, but also in Elzinga (2010) sense -->

Complexity in the family domain has been defined as the combination of family structure and sibling composition during childhood [@manning_family_2014]. Where a "simple" family may be defined as a household with both parents present and a few number of full-siblings, a complex family may consists of step-parents as well as half- and step-siblings. The four main demographic drivers of complexity are birth to lone mothers, separation and divorce, repartnering, and multi-partner fertility [@thomson_family_2014]. Phenomena which have all become more salient during the Second Demographic Transition. Although, it is primarily driven by the adults involved it clearly has consequences for the family life of children. The variety of childhood family structures have been extensively documented in the US context, but less so in Europe. I will first summarize these rich findings from the US, then turn toward the European context and finally discuss some of the literature on the consequences of childhood family complexity. 

Using data from the Survey of Income and Programme Participation (SIPP), @manning_family_2014 describe changes in family complexity for US children over the period 1996-2009. The share of children who at the time of interview were living in a complex household --- defined as living with either a step-parent, single parent, or full- or half-sibling --- increased from 30.8 to 33.6%. The authors found that family complexity was especially prevalent in families where neither parent had a college degree as well as for Black children. However, the steepest increase in complexity during the observation period was found among Hispanic children. Further analysis of the SIPP showed that in 2008, the income-to-needs ratio for children in complex families was 2.0, contrasted to 3.0 in simple families [@brown_family_2015]. Further, 28.3% of complex families received public assistance (compared to 17.4% of simple families), on average. Similar ethnic differences were found by @blau_demographic_2008, where --- in the 1979 cohort of the National Longitudinal Survey of Youth --- children of Black mothers spent on average 34.1% of their childhood living with both their biological mother and father, compared with 72.8% for whites and 64.1% for Hispanics. Children to Black mothers also had, on average, more stepfathers than other groups. Accounting for historical changes in family structure, @fomby_continuity_2022 found that complexity increased during the late 20^th^ century but arrived at a state of "stable complexity" in the 2010s, where household composition also became increasingly similar between non-Hispanic Black and White children. 

In Europe, similar patterns have been observed where children from socio-economically disadvantaged groups experience family disruptions to a larger degree. In Sweden, children are more likely to experience parental separation if their parents are less-educated [@kennedy_childrens_2010]. This is also echoed in @kalmijn_new_2021, where similar results were found also for Belgium, Czech Republic, Estonia, France, Hungary, Norway and the United Kingdom. They also found the this differential risk of parental divorce has been increasing since the 1980's. When it comes to stepfamily formation, @turunen_entering_2011 found no effect of parental education on the risk of entering a stepfamily after having lived with a single parent (either due to separation or being borne to a lone parent). Thus, the risk of entering a complex family seems to be largely dependent on the different separation risks that children experience.

It is important to note that there are large cross-national variations in risks of parental divorce and stepfamily formation in Europe. Using life-table methods, @andersson_life-table_2017 describe these risks for a large set of European countries. From their descriptives we can see that the risk of parental separation by age 15 is highest in parts of Eastern and Northern Europe (Russia 42%, Estonia 36%, Lithuania 35% but also France 35%) while its the lowest in Southern Europe and other parts of Eastern Europe (Georgia 10%, Bulgaria 12%, Spain 12% and Italy 12%). There are no clear patterns when it comes to the proportions experiencing a stepfamily after parental separation but it varies from 14% (Georgia and Italy) up to 60% (Belgium).

Although there are no direct comparative studies on the occurrence of half-siblings in different European countries, we can at least infer trends from adult data on differences in multipartner fertility. @thomson_childbearing_2021 finds that for mothers, multipartner fertility is most common in Czech Republic and Estonia where 22.8 and 21.7% of mothers with multiple children have had their children with two or more partners. These countries are then followed by France, Lithuania, Norway and Sweden where the number is around 16% of mothers with two or more children. The lowest occurrence of multipartner fertility is found in Georgia, Spain and Bulgaria. Thus we see that there is some overlap with the patterns found by @andersson_life-table_2017, half-siblings are probably more common in the Nordic and Baltic states, along with some nations in Central Europe. While it is more uncommon in Southern Europe and parts of Eastern Europe. This is an important area of research which deserves more attention and it is of particular interest to understand how present half-siblings are in each others lives. Findings from Sweden suggest that the time lived with similarly aged half-siblings is substantially less than the time lived with full siblings [@andersson_oh_2020].

Other aspects which are important to consider when tracing children's family experiences, both in the US and in Europe, is if cohabitation is included as a measure as well as the prevalence of joint physical custody (or shared residence). Relying on the US National Survey of Family Growth,  @rally_cohabitation_2004 found that when adding family transitions into and out of cohabitation, family instability (calculated from transitions into and out of parental partnerships) increased by 30% for White children and over 100% for Black children. This highlights the importance of including cohabitation in studies on family complexity. For all countries studied, @andersson_life-table_2017 also found that experiencing a parental separation is more common for children born into cohabiting unions.

Further complicating research on children's family structure over the life course is the phenomenon of joint physical custody. In a majority of European countries, joint physical custody is a fringe arrangement that only 5% or less of children experience, whilst it is quite common in Sweden (21%) and Belgium (13.5%) [@steinbach_joint_2021]. This makes it hard to examine the actual family experiences of children based on surveys where parents are the anchors or where children's experiences are not tracked after separation on both the maternal and paternal side. If the prevalence was similar across countries it would be simpler to approximate children's experiences based on parental reports. Current praxis of following children's family trajectories on the maternal side will now under report the total experienced complexity more in countries such as Sweden and Belgium whereas it will be a good approximation in countries where joint physical custody is uncommon. For instance, estimates of the proportion of children with half-siblings have been found to be lower if there is fertility information only on one parent [@cancian_evolution_2011]. There are findings from the UK which show that children's family complexity (measured with a compound complexity index) are similar between father's and mother's family lives over childhood [@rowold_cohort_2022], indicating that a fair share of complexity is missed when only accounting for children's experiences on the maternal side. Attempts to ameliorate this issue have been made [see @kalmijn_multi-actor_2018] but is so far uncommon. 

Family instability and changes in household composition have further been linked to a diverse range of outcomes such as higher levels of behavioral problems for children born to lone mothers and those whose mother's gave birth to half-siblings [@fomby_family_2017], higher risk of bullying if living in a single parent household [@laftman_national_2017], excess externalizing behavior for boys in stepfamilies [@mostafa_impact_2018] as well as decreased psychological well-being and educational attainment of children experiencing parental divorce [@gahler_parental_2015].

In a review of the research on family diversity and complexity among adolescents, [@pearce_increasing_2018] argues that to better understand both the family circumstances of adolescents as well as how family characteristics are related to certain outcomes, we are in need of new methods. They suggest, among others, the use of "profile based" methods, such as latent class analysis and cluster analysis. So far, very few studies have been conducted on children's family structure using this type of methodology but we will look more closely at two studies that have. 

Using the U.S. National Longitudinal Survey of Youth, @mitchell_pathways_2013 created a five type classification of children's family structure on the maternal side using latent class analysis. The model distinguished between living with (1) married biological parents, (2) cohabiting biological parents,  (3) cohabiting stepfather, (4) married stepfather and (5) single mother. The five class solution created typologies that Mitchell named, "Married biological parents who divorce", "Long-term single mothers", "Continuously married biological parents", "Cohabiting biological parents who marry or break up" and "Gain a stepfather". This study can be credited as an early attempt at finding childhood family trajectories but are in some senses lacking. One of the largest drawback is making a distinction between cohabiting parents and married parents. Although this might be important from an adult perspective, it probably makes no difference in how children perceive the parental relationship. It further gives rise to the category "Cohabiting biological parents who marry or break up", which really makes no substantive sense. Classifying children who experience a separation into the same group as those who might experience a strengthening of the parental relationship seems to be just an artifact of how the model and it's states are constructed. Therefore, we need to be aware of how model construction can influence the outcomes with this type of data-centric analysis. 

A more recent study, using the children of the participants in the National Longitudinal Survey of Youth, performed and compared classifications of family trajectories through both latent class analysis and sequence analysis [@johnston_two_2020]. This study did not distinguish between cohabitation and marriage and revealed some further patterns in family structure changes, and where each pattern was intuitively interpretable. In particular, the latent class analysis distinguished classes on the basis of timing, where early and late entry into both a family with a sole mother as well as into a stepfamily became distinct types. It was also found that the different types identified both by sequence analysis and latent class analysis predicted the risk of later adolescent alcohol consumption. 

__TO DISCUSS__

Explanations: @cavanagh_family_2019 @fomby_family_2017 @cavanagh_family_2008 @fransson_psychological_2016

_TRANSITION FROM FAMILY PROFILES AND ROLE OF EVENT TIMING TO LIFE COURSE THEORY_

## Life course theory 

According to Elder et al. [-@elder_emergence_2003], life course theory came about from a "desire to understand social pathways, their developmental effect, and their relation to personal and social-historical conditions". Since its incipience in the 1970's, the *life course* has grown from a scientific idea to a theoretical framework and even a research field in its own right. With the launch of *Advances in Life Course Research* as a quarterly journal, it was declared that the life course is coming of age [@billari_life_2009]. 

The main theoretical concepts from the mature life course research which will be used in this thesis is those of *transitions*, *trajectories*,  *pathways* and *linked lives*. *Transitions* are state changes that are discrete and time-bound [@george_sociological_1993]. These can occur in multiple domains --- such as the social, biological, or economic --- and examples include divorce, menopause or getting a driver's license. In turn, multiple transitions make up life *trajectories*. These consists of longer periods of change and stability, and although marked by a series of transitions they have a continuity of direction [@hutchison_life_2005]. A family formation trajectory by Hollywood blueprint would consist of dating followed by cohabitation and finally a wedding, marking a transition into married life. 

Just as trajectories are made up of a set of transitions, *pathways* are constituted by interlocking transitions and trajectories across time and domains [@macmillan_families_2005]. In particular, pathways are often in relation to entering and exiting social institutions such as a career or family. Although the combinatorial possibilities of transitions and trajectories are vast, the number of actualized pathways are usually limited and in relation to normative systems [@white_application_2014]. A way of thinking about pathways is that they are lived in relation to schema; that is, blueprints of culturally appropriate behavior, which are enacted more or less consciously [@sewell_theory_1992]. Schema inform individual heuristics for the temporal organization of the life course by providing a set of rules and models for how to link pathways over time [@macmillan_families_2005]. A modern Scandinavian pathway in relation to family formation would be forming a cohabiting union followed by having children, skipping the previously normative transition of marriage [@hofacker_patterns_2014]. 

Although often emphasized in the literature, the power of normative schema in shaping the individual life course has been contested [@macmillan_structure_2005]. Apart from norms, institutional and structural processes also shape lives and are able to produce homogeneity in life patterns [@marini_age_1984], as we have also seen in the discussion on the proliferation of the SDT. A further way of thinking about pathways, uncommitted to theoretical assumptions about their generation, is simply as a set of probabilities of transitioning between different states. But rather than a set of independent probabilities, they are path dependent with a certain degree of historicity, which can give rise to common patterns in a population [@rutter_pathways_1989]. 

Finally, a central idea in the life course framework is that of *linked lives*. This concept highlights the interdependence of our lives and that transitions that occur in one persons life often entail transitions for other people [@elder_emergence_2003]. This principle has particularly been applied in studies on the family and intergenerational relations [@carvalho_family_2021]. Already in one of the first studies using a life course perspective, @elder_children_1974 found that children's development could be impeded by the economic hardships experienced by their parents. This interdependence between parents and children, and how their social lives are intertwined, has since been explored in much depth and breadth. 

It is with this interdependence in mind, that I approach my object of study. Children's early family experiences are highly dependent on their parents' partnership trajectories, both in terms of vertical and horizontal family relations. At the same time, children's early family trajectories may themselves be associated with future outcomes. Thus, it is of interest to know how children's family trajectories are constituted. How to conceptualize and operationalize these trajectories is what I turn to in the next section. 


## Operationalizing family complexity over the life course

As has been outlined, the family structure of children may exhibit many different patterns and if we want to fully understand this complexity also within a life course framework, we need to study how these patterns play out over time. One fruitful analytic method which has been used within the life course literature is sequence analysis [SA, for an overview see @liao_sequence_2022]. The foundation of SA lays in describing temporal processes as a series of states and state changes over discrete time units (such as months or years). It thereby provides a format for describing individual life course trajectories.

A usual approach for analyzing this type of data is through the use of cluster analysis [@piccarreta_holistic_2019]. This is a data-centric approach which deviates from traditional statistical methodologies in two main ways: 1) the units of analyses and 2) data assumptions. @aisenbrey_new_2010 show these differences by contrasting event history analysis with SA. Where event history analysis studies probabilities of transitions, SA studies patterns in trajectories as their main units of analysis. When it comes to assumption about the data, event history analysis assumes that the data is generated by a stochastic process whereas the algorithmic clustering makes no assumptions about the underlying process. This also leads to differences in practice, where event history analysis usually employs models where included covariates are derived from theory or previous empirical research, SA relies on algorithmic clustering of the data to identify groups of interest. These clusters are then used in two main ways: most commonly they are used as a dependent categorical variable in multinomial regression to find how other covariates predict life trajectories [@piccarreta_holistic_2019] but the clusters can also be used as predictor variables to see how life trajectories influence later life outcomes [e.g. @johnston_two_2020].

However, a fair amount of critique has been leveled against the clustering techniques employed by sequence analysis. To put this critique in methodological context, I will first outline the general steps in producing sequence analytic clusters before discussing the methodological drawbacks. The first step is to create a distance (or dissimilarity) matrix by calculating the distance between each pairwise combination of individual sequences in the data. Since the seminal work of @abbott_optimal_1986, the Optimal Matching algorithm has provided a standard way of calculating such a distance metric. The basic intuition is that we can permute a sequence by two kinds of operations to arrive at another sequence. The operations are substitutions (replacing one state by another) and insertions/deletions of states (commonly known as _indels_). In most cases, it is however not as straightforward as just summing up the number of operations needed to permute one sequence into another to get their distance. As the probability of transitions between states differ between pairs of states, or between substitution and indel operations, we need a substitution matrix which gives a cost to each possible substitution between pairs of states. We also need to assign a cost to indel operations. Thus, the optimal matching algorithm tries to find a set of operations that optimize the cost of permuting one sequence to another. This optimized cost is used as a measure of distance. Using the finalized distance matrix, it is possible to apply a clustering algorithm to derive a set of clusters, this is commonly done through Ward's method [@aisenbrey_new_2010].

The criticism leveled towards this methods is focused on two main areas: firstly, on the perceived lack of substantial results produced by the methodology and secondly, and perhaps more importantly, on the relationship between methodological assumptions and social processes. Both @wu_comments_2000 and @levine_but_2000 criticize SA for not delivering on early promises and that substantial results and new areas of research have not been borne out of SA. As summarized in @liao_sequence_2022, there is a large variety of sociological articles being published using SA which would render that point of contention moot, but it does emphasis the importance of scrutinizing the SA methodology and if it is a valid tool for modeling or describing social processes.

@wu_comments_2000 offers a critique of SA where the most pressing is if a discrete sequence with events coupled with substitution and indel costs is a good model of life trajectories. The main problem lies in determining costs, as each cost is an assumption the researcher has to make, and that costs in the optimal matching algorithm are symmetric. This symmetric assumption means that transition from state $s_j \rightarrow s_k$ carries the same cost as transition from $s_k \rightarrow s_j$. But the question posed by Wu, is if the transitions between social states, such as employment and unemployment, are symmetrical in their probabilities or not. Which would imply that sequence representation through optimal matching is not isomorphic to the social processes they purport to represent. Further, Wu argues that as determining substitution and indel costs is up to the discretion of the researcher, it introduces more assumptions and potential biases in the analysis than proponents of SA had up to then acknowledged. This is a valid criticism but loses some of its relevance if we consider S as an exploratory or descriptive analysis method, rather than a substitute to statistical modeling.

__TO DISCUSS__
Problems with clustering: @warren_different_2015 @han_comparing_2017 @von_luxburg_clustering_2010

# Data and Methods

## Analytical strategy

Before describing the data material and the details of the analysis, I will outline the analytic strategy of this paper and how it relates to the research question, previous studies and the strengths and weaknesses of SA. The first step consists of defining family states and generating individual family sequences. As the data does not track both original parents of a child, I will only follow the mother's partnership trajectory with the assumption that children are residing at least part time with their mother during their childhood. Thus, join physical custody is not taken into account, and the family trajectories on the father's side are not measured. This will make the measures of family complexity biased downwards. A positive note is that it will avoid the problem that men tend to under report their number of children in retrospective surveys [@juby_where_1999]. So although an analysis solely on the mother's side is a limitation, we can proceed as long as we keep that limitation in mind. Having constructed the parental structure sequence, I compute sequence dissimilarities and generate clusters for the full study population. The final number of clusters are selected using both cluster quality criteria and the sociological significance of the cluster division.

The problems of the validity of clustering typologies have been recognized for a long time, and methods for ameliorating this deficit have been developed within the bioinformatics literature [for an overview see @von_luxburg_clustering_2010], few SA studies within the social sciences have taken this issue to heart. In order to improve the validity of the typology, I will use a parametric bootstrap technique proposed by @hennig_flexible_2015 and implemented for sequence data by @studer_validating_2021 to aid with the selection of the final clusters.

Once the family typology is in place, I will show the proportion of the different family types for each region in the study, thus providing a descriptive of how children's family trajectories vary across Europe. I will also show how the different typologies are associated with having full- and maternal half-siblings, on a regional basis. Apart from mapping the relationship between family trajectories and sibling composition, it will also be one of the first cross-European analyses on half-sibling prevalence. This will further contribute to the understanding of the relationship between parental union history and sibling composition as well as showing the differences and similarities between European regions.

## Data

The data used in this study comes from the birth and union histories in the Harmonized Histories data set [@perelli-harris_harmonized_2010]. This data set is compiled from a set of national surveys, primarily the Generations and Gender Survey, but also with other nationally representative surveys on partnerships and childbearing. The data is based on individual retrospective reports on partnership and childbearing histories.

I start of by excluding several countries from the data set. Germany is excluded due to its poor family history data and the UK is excluded due to oversampling respondents with low socioeconomic status. Uruguay, Kazakhstan, Canada and the U.S are removed as they are not European countries.

I then have __xx__ total observations from 16 different countries. I remove all men from the sample. Next, I remove childless women and respondents with missing birth information. I remove those where there is missing information on time of union or dissolution, overlapping relationships and those where a union dissolution is reported to have occurred before the union start.

In order to be able to analyze the trajectories from the children's lived experience I reshape the data so that each child becomes a unit of observation. At the start I have __xx__ children. I remove children with missing birth month or where the birth is reported to have occurred before the mother is born. As I want to track children's' experiences until the end of age 15, I remove those in the sample who have not turned 16 at the time of interview as well as those who left home or died before their 16^th^ birthday. In order to capture the experiences of children after the advent of the SDT, I remove all children born before 1980. In total 48,163 children remained in the sample. For a visual overview of the data funnel at each step, see Figure __x__ in the Appendix. 

## Sequence generation and sibling composition

For each child a family history was constructed on the maternal side, which is composed of the mother's partner history during the child's first 15 years of life. In order to construct this history, a STate-Sequence format [STS, @gabadinho_analyzing_2011] is used. This means that for each month of the child's life a column is generated which is filled with values describing the mother's union status and the child's number of siblings. The process time for analysis is from birth until turning 16, with quarterly precision. In STS format this is represented by 60 consecutive states, each capturing a three month period. 

Firstly, union spells were constructed. As there is no information on who is the father of the child if a mother was partnered at time of birth, that partner was assumed to be the other main parent. Then for each quarter, the representative state was coded as either *Single*, *Original parents* or *Step* depending on the mother's partner status in the first month of each quarter. The state is coded as *Single* if the mother is not in a union during the first month of the quarter. It is coded as *Original parents* if the mother is still in a relationship with the same partner as at the time of birth. Finally, a state is coded as *Step* if the mother is with a different partner than at time of birth. All consecutive partners are assumed to be new, i.e. no re-partnering with a former partner is assumed to occur. I do not distinguish between opposite-sex and same-sex partnerships, and the original parents are not necessarily the same as the biological parents, but simply constitute the partnership that the child is born into. Formally, the maternal union sequence can be described as the vector $\textbf{S}$ where  

$$
\textbf{S} = (s_1,s_2,s_3, \cdots, s_{60}) | s_i \in \{\textrm{Original parents, Single, Step} \}
$${#eq-sequence-example}

An example sequence of a child's first two years of life, where the mother is first partnered for 12 months (four quarters), then single for 6 (two quarters) and finally repartnered for 6 months (two quarters) is visible in @tbl-state-ex.

```{r}
#| echo: false
#| results: "asis"
#| label: tbl-state-ex
#| tbl-cap: "Two year union state sequence example"

months <- paste0("", 1:8)
states <- c(
  "OP", "OP", "OP", "OP",
  "Si", "Si", "St", "St"
)
knitr::kable(
  rbind(
    months,
    states
  ),
  "simple",
  row.names = FALSE
)
```

To determine sibling composition I use a simple binary indicator if each child has at least one full-sibling and if each child has at least one half-sibling. Full-siblings are defined as siblings born to the same mother as well as being born in the period when the mother is cohabiting with the other original parent of the anchor child. A half-sibling is a sibling which is born either in a period of stepfamily for the anchor child or if the anchor child or sibling is born during a state of maternal single hood.

## Analysis

Having constructed the maternal union sequence, I calculate a dissimilarity matrix using an Dynamic Hamming Distance algorithm. This algorithm puts more restrictions on the data than Optimal Matching by requiring that the sequences are of the same lengths. This also allows the Dynamic Hamming Distance algorithm to calculate dissimilarities without using indel operations. One of the benefits of this approach is that indels warp time in the data, and thus reduces sensitivity to sequence timing [@lesnard_setting_2010]. Another benefit is that indel costs needn't be defined which reduces the amount of decisions, and potentially biases, when generating the dissimilarity matrix. The substitution costs are derived from the transition rates in the data and thus reflects transition probabilities in this particular sample. This is the most data-centric approach possible for defining substitution costs and also reduces potential researcher biases, with the drawback of potentially reducing the external validity of the final clustering solution. 

This dissimilarity matrix is used to cluster the sequences, which is done using Ward's method. This is an agglomerative hierarchical clustering algorithm that produces cluster by minimizing the within-cluster variance, using the squared dissimilarity distances as inputs [@murtagh_wards_2014]. Being an agglomerative method, it starts with $k$ number of clusters where each observation in the sample is its own cluster, thus $k = N$ in the first step. The algorithm then progressively merges clusters until there is only one cluster containing all $N$ observations. Larger clusters are formed in steps by merging the two clusters with the shortest distance, where distance is defined by the increase in sum-of-squares if the clusters are merged. Given that this clustering method produces solution between $1$ and $N$ clusters, a choice regarding the final number of clusters needs to be made. This choice will be based on sociological relevance and cluster quality indices.

In choosing the number of clusters, it is of interest to maximize internal cluster homogeneity as well as between-cluster differences, and a common measure of this propriety is the Average Silhouette Width , introduced by @kaufman_finding_1990. This measure produces a value between -1 and 1, where a larger value indicates a stronger structure. I also use Hubert's C , which captures the discrepancy between the partition obtained and the theoretically best partition, given the particular number of groups and their distances [@hubert_general_1976]. Hubert's C is defined in the range $[0,1]$ and a smaller value indicates a better partitioning. The issue with assessing cluster quality is that, unlike in traditional statistical modelling, there is no null model to compare the results with. So, even though @kaufman_finding_1990 propose an interpretation of the Average Silhouette Width measure, it cannot be generalized to all cluster applications. 

A potential solution is to use parametric bootstrapping to generate _null-model like_ distributions, cluster them, calculate a cluster quality index and compare it to the quality index value of the clusters from the actual data [@hennig_flexible_2015]. This has recently been applied to SA [@studer_validating_2021] and to my knowledge, this study will be one of the first to apply this technique. In practice, I generate a data set of 50 sequence sets that are similar but non-clustered using a combination of both sequencing and timing parameters [for the detail of the process see @studer_validating_2021]. Then the Average Silhouette Width and Hubert's C are calculated for each cluster solution ranging from 2 to 10 clusters for each of the 50 bootstraps. This produces a confidence interval for each of the quality measures for each $k$ grouping. This provides a baseline measure of the quality indices for data that is not clustered. If the index values of the cluster solutions for the actual data falls out of this bootstrapped range, it is a strong indicator that the particular clustering solution represents a "true" cluster and not just an artifact of the clustering process. However, that a cluster is true from a purely data-centric view, does not mean that it holds any form of social significance. Therefore, apart from using cluster quality indices, I will also look at the theoretical basis of the best performing cluster solutions when selecting the number of clusters to use. Further, I only consider 2 to 10 clusters as a larger number of cluster would be both unwieldy in terms of discussing the number of groups and the number of members in the smaller clusters would be too few to carry out any form of description on the country level. 

I provide a visual summary of the sequence clusters by a set of three plots for each cluster. First, a state distribution plot which shows the relative frequency of each state in the population by month. Second, a modal plot which shows the relative frequency of the most common state for each time point. Third, a medoid plot which shows the sequence with the smallest average dissimilarity to all other sequences in the group. It can be seen as a "median" sequence, or as the most typical trajectory within the cluster group. I also label the clusters with qualitative descriptions that aim to capture the general family experience of each cluster. 

In the next step, I provide a description of the proportions of children which belong to each cluster, on a country level. This allows us to detect similarities and differences between European countries as it relates to children's family experiences. In the final step I also relate parental union histories to sibling composition by cross-tabulating cluster membership with indicators of having full- and half-siblings for each country. 

The analysis is performed in R and relies on the packages **TraMineR** and  **WeightedCluster** [@studer_weightedcluster_2021] for the sequence and cluster analysis. 

## Ethical considerations

The Harmonized Histories data set is de-identified to protect the personal integrity of the study participants and all analysis presented are as aggregated results. In order to get access to the data I have signed a pledge of confidentiality, pledging not the share the data and only perform analysis related to the approved project. The data collection has been approved in each individual country and has been done under the oversight of the GGP Ethics Board and its Data Protection Office. 

In order for the research to be both transparent and reproducible, all code for generating the results, as well as the thesis, is available as a reproduction package at GitHub [@gustafsson_addegezonfamily_complexity_in_childhood_2022].


# Results

## Descriptives
 @tbl-1 shows basic descriptives of the study population.

```{r table1}
#| echo: false
#| results: "asis"
#| label: tbl-1
#| tbl-cap: "Descriptive information on survey years, cohorts included and number of children from each country"

tar_read(table_1)
```

## Cluster diagnostics

```{r}
#| echo: false
#| label: fig-cwi
#| fig-cap: "Standardized Average Silhouette Width and Hubert's C for actual and bootstrapped data"

#tar_read(p_cluster_cqi)
```


```{r}
#| echo: false
#| results: "asis"
#| label: tbl-proportion-family-type
#| tbl-cap: "The proportion of children classified according to each family type by European region."

#tar_read(tab_proportion_family_type) |> 
#  kable_styling(font_size = 8)
```


## Final clusters

```{r}
#| echo: false
#| label: fig-joint-plot
#| fig-cap: "Plot of the five cluster solution"
#| fig-height: 8

#tar_read(p_clusters_5)
```

## Sibling composition

# Discussion

## Limitations

# Conclusion

# Acknowledgments

# References

::: {#refs}
:::

# Appendix