---
title: "Childhood Family Trajectories in 16 European Countries"
subtitle: "WORK IN PROGRESS"
author: "Andreas Gustafsson"
date: today
date-format: long
format: 
  pdf:
    documentclass: article
    papersize: a4
    geometry: "left=3cm, right=5cm, bottom=3cm"
    include-in-header:
      text: |
        \usepackage{booktabs} 
        \usepackage{xcolor} 
        \usepackage{siunitx}
        \usepackage{lscape}
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
        \usepackage{geometry}
        \usepackage{fontspec}
        \usepackage{ragged2e}
        \usepackage[british]{babel}
        \usepackage{setspace}
        \usepackage{titling}
        \usepackage{etoolbox}
        \AtBeginEnvironment{table}{\sffamily}
        \pretitle{\begin{flushleft}\Large}
        \posttitle{\end{flushleft}}
        \preauthor{\begin{flushleft}}
        \postauthor{\end{flushleft}}
        \predate{\begin{flushleft}}
        \postdate{\end{flushleft}}
    include-before-body:
      text: |
        \onehalfspacing
        \RaggedRight
        \hypersetup{linkcolor=black}
        \tableofcontents
        \newpage
        \listoftables
        \listoffigures
        \hypersetup{linkcolor=DarkOrchid4}
    latex-engine: xelatex
link-citations: true
urlcolor: blue
reference-location: block
bibliography: "bibliography.bib"
---

```{r}
#| include: false
#| warning: false
#| echo: false

# include library and targets source
library(targets)
library(tarchetypes)
library(fst)
library(ggplot2)
library(forcats)
library(data.table)
library(kableExtra)
tar_config_set(store = "../_targets/")
knitr::opts_chunk$set(
  out.width="100%",
  dev="png"
)

knitr::knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",")
})

```

# Introduction

The European family dynamics of the late 20th and early 21st century has undergone large changes as compared to previous decades. In particular through increasing family complexity. From the amassed family demography literature, the majority of attention has been placed on investigating the family complexity generating processes of divorce, re-partnering and multipartner fertility from the adult perspective. Since it has been proposed that family transitions have a negative effect on the well-being of children [@fomby_family_2007] and because there is evidence of this effect [@hadfield_changes_2018], it is important to understand to what extent children experience family transitions and what the patterns are. Demographic studies that have been conducted on children's experience of family disruption and parental re-partnering [such as @andersson_childrens_2002;@kennedy_childrens_2010; @thomson_family_2014] have primarily looked at the degree to which children experience particular events. Albeit important, this does not fully capture the accumulated experience of family complexity over childhood. 

Further, there has been ample research on how these family demographic changes have developed in Europe, and how the processes have differed between European regions. This has often been approached through the Second Demographic Transition theory. Despite this rich flora of comparative research, very little has been done to document the implications for children's family trajectories and to what degree they differ between countries. 

The objective of this thesis is to provide a descriptive analysis of children's family structure trajectories in Europe. In particular I will answer the questions (1) What are the most common family trajectories among European children, and how do they vary across European countries and regions? (2) How does the probability of experiencing a particular family trajectory differ by the mother's socioeconomic status, as measured by level of education? This will be accomplished by generating a set of family history sequences for European children, based on the Harmonized History data set. Analysis will be performed by sequence analysis, through which I will generate clusters of the most common family trajectories of European children. I will then perform a descriptive analysis of the proportion of children who experience these trajectories on a country by country basis. Further, I will describe how the most common trajectories are associated with the mother's socioeconomic status. 

Before describing the methods in greater detail, I will outline the main theoretical frameworks of this thesis as well as previous research on children's experiences of family structure changes. In the end I will present the results of my analysis and put them into context in the discussion and conclusion sections.


# Theory and previous research 

## The consequences of family complexity


Complexity in the family domain has been defined as the combination of family structure and sibling composition during childhood [@manning_family_2014]. Where a "simple" family may be defined as a household with both parents present and a few number of full-siblings, a complex family may consists of step-parents as well as half- and step-siblings. The four main demographic drivers of complexity are birth to lone mothers, separation and divorce, re-partnering, and multi-partner fertility [@thomson_family_2014]. Phenomena which have all become more salient during the latter 20^th^ century. Understanding how family structures change during childhood is important because family transitions and different family arrangements have tangible consequences for the children involved. In this section I will outline what some of these consequences are, focusing both on the consequences of experiencing particular events as well as living in certain family arrangements.

One of the most well studied childhood family transition is divorce, along with its consequences on proximate and future child outcomes. These consequences occur in multiple domains, ranging from well-being to educational outcomes. @averdijk_parental_2012 found that parental separation lead to an increase in both aggressive and internalizing behavior among children in Zurich, controlling for both parental conflict, maternal depression and financial difficulties. The effect was partly mediated by maternal depression. Similar results were found in the U.S context, where children of divorce, who were propensity-score-matched to children of intact families, showed higher levels of both internalizing and externalizing behavior following parental separation [@weaver_mediation_2015]. This effect was found both in the short- and long-term following separation. There are also potential health consequences of divorce, where German children experiencing divorce, studied using a panel setup, were found to increase their BMI after divorce [@brockmann_unhealthy_2013]

There are also consequences on school outcomes. @havermans_family_2014 found an association between divorce and less school engagement among Belgian children. Large part of the effect was mediated by the parent-child relationship, financial problems at home and parental conflict. Swedish children experiencing divorce have also been found to show decreased psychological well-being and lower educational attainments [@gahler_parental_2015]

In opposition to divorce, the event of re-partnerning has also been studied, but not to the same degree. @bachman_is_2009 studied adolescent from low-income urban US families, and found that neither risks nor benefits increased as a result of maternal marriage on a wide battery of adolescent well-being indicators. Although introducing a new partner in the household increased financial resources, it was not associated with improvements in home-environment or maternal well-being, which could explain why adolescent functioning did not increase significantly. In a further US study, children's academic achievement was not found to be improved when their single mother married [@wagmiller_does_2010]. The small benefits to academic performance was also to a large degree distributed among the children to mother's with higher education. There is yet little evidence for the case that maternal re-marriages increases the life chances and well-being of their children [@hadfield_changes_2018]

Apart from the outcomes of family transitions, there is also a literature on the effects of the long term states of living with a single mother or in a stepfamily, contrasted with living in a nuclear family with original parents. For the case of single mother families in the U.S, @fomby_family_2017 found that children born to lone mothers showed higher levels of behavioral problems than those born into nuclear families. Youth in single mother families in Germany have also been found to experience lower well-being than those in living with their mother and father, or their mother and a stepfather [@walper_effects_2015]. There are also external factors to consider, which might influence the well-being of children in single mother families, one example comes from Sweden where living in a single parent household has been found to be associated with a higher risk of bullying for children [@laftman_national_2017].

Finally, there have also been seen that living in a stepfamily may be associated with negative outcomes for children... __TO BE WRITTEN__


<!-- excess externalizing behavior for boys in stepfamilies [@mostafa_impact_2018]  -->


All in all, these findings suggest that both family transitions and long-term family states matter for the well-being of children in several dimensions. A consequence of this is also that studies that focus on either transitions or states will miss out the other process, indicating a need for a more holistic approach to the study of children's family lives. This has also been highlighted by @pearce_increasing_2018, who argue that to go further in our understanding of the associations between family complexity and its related outcomes, we are in need of new methods. A set of method which they suggest as a potential way forward are "profile-based" methods, such as latent class analysis and sequence analysis. A more detailed discussion of this approach will follow in later sections.


## Tracing family trajectories

The variety of childhood family structures and transition have been documented to some extent in Europe from a child-centered perspective. A general patterns that has been found is that children from socio-economically disadvantaged groups experience family disruptions to a larger degree. In Sweden, children are more likely to experience parental separation if their parents are less-educated [@kennedy_childrens_2010]. This is also echoed in @kalmijn_new_2021, where similar results were found also for Belgium, Czech Republic, Estonia, France, Hungary, Norway and the United Kingdom. They also found the this differential risk of parental divorce has been increasing since the 1980's. When it comes to stepfamily formation, @turunen_entering_2011 found no effect of parental education on the risk of entering a stepfamily after having lived with a single parent (either due to separation or being borne to a lone parent) for Swedish children. Thus, the risk of entering a complex family seems to be largely dependent on the different separation risks that children experience.

It is important to note that there are large cross-national variations in risks of parental divorce and stepfamily formation in Europe. Using life-table methods, @andersson_life-table_2017 describe these risks for a large set of European countries. From their descriptive analysis we can see that the risk of parental separation by age 15 is highest in parts of the post-Socialist countries (Russia 42%, Estonia 36%, Lithuania 35% but also France 35%) while its the lowest in Southern Europe and the southern parts of post-Socialist Europe (Georgia 10%, Bulgaria 12%, Spain 12% and Italy 12%). There are no clear patterns when it comes to the proportions experiencing a stepfamily after parental separation but it varies from 14% (Georgia and Italy) up to 60% (Belgium). As for births to single mothers, who are neither in cohabitation or marriage, The Czech Republic and Russia stands out with 13% of children experiencing this. But there are also high levels in France, Estonia, Lithuania and Poland (8% of all children in these countries.) On the other side of the spectrum we have the Sweden (2%), Norway (3%), Netherlands (3%) and Georgia (3%). 

Other aspects which are important to consider when tracing children's family experiences, both in the US and in Europe, is if cohabitation is included as a measure as well as the prevalence of joint physical custody and father-only households. Relying on the US National Survey of Family Growth,  @rally_cohabitation_2004 found that when adding family transitions into and out of cohabitation, family instability (calculated from transitions into and out of parental partnerships) increased by 30% for White children and over 100% for Black children. This highlights the importance of including cohabitation in studies on family complexity. For all countries studied, @andersson_life-table_2017 also found that experiencing a parental separation is more common for children born into cohabiting unions. As for joint physical custody, in a majority of European countries it is a fringe arrangement that only 5% or less of children experience, whilst it is quite common in Sweden (21%) and Belgium (13.5%) [@steinbach_joint_2021]. This makes it hard to examine the actual family experiences of children based on surveys where parents are the anchors or where children's experiences are not tracked after separation on both the maternal and paternal side. If the prevalence was similar across countries it would be simpler to approximate children's experiences based on parental reports. Current praxis of following children's family trajectories on the maternal side will now under report the total experienced complexity more in countries such as Sweden and Belgium whereas it will be a good approximation in countries where joint physical custody is uncommon. For instance, estimates of the proportion of children with half-siblings have been found to be lower if there is fertility information only on one parent [@cancian_evolution_2011]. There are findings from the UK which show that children's family complexity (measured with a compound complexity index) are similar between father's and mother's family lives over childhood [@rowold_cohort_2022], indicating that a fair share of complexity is missed when only accounting for children's experiences on the maternal side. Attempts to ameliorate this issue have been made [see @kalmijn_multi-actor_2018] but is so far uncommon. 

In a review of the research on family diversity and complexity among adolescents, [@pearce_increasing_2018] argues that to better understand both the family circumstances of adolescents as well as how family characteristics are related to certain outcomes, we are in need of new methods. They suggest, among others, the use of "profile based" methods, such as latent class analysis and cluster analysis. So far, very few studies have been conducted on children's family structure using this type of methodology but we will look more closely at two studies that have. 

Using the U.S. National Longitudinal Survey of Youth, @mitchell_pathways_2013 created a five type classification of children's family structure on the maternal side using latent class analysis. The model distinguished between living with (1) married biological parents, (2) cohabiting biological parents,  (3) cohabiting stepfather, (4) married stepfather and (5) single mother. The five class solution created typologies that Mitchell named, "Married biological parents who divorce", "Long-term single mothers", "Continuously married biological parents", "Cohabiting biological parents who marry or break up" and "Gain a stepfather". This study can be credited as an early attempt at finding childhood family trajectories but are in some senses lacking. One of the largest drawback is making a distinction between cohabiting parents and married parents. Although this might be important from an adult perspective, it probably makes no difference in how children perceive the parental relationship. It further gives rise to the category "Cohabiting biological parents who marry or break up", which really makes no substantive sense. Classifying children who experience a separation into the same group as those who might experience a strengthening of the parental relationship seems to be just an artifact of how the model and it's states are constructed. Therefore, we need to be aware of how model construction can influence the outcomes with this type of data-centric analysis. 

A more recent study, using the children of the participants in the National Longitudinal Survey of Youth, performed and compared classifications of family trajectories through both latent class analysis and sequence analysis [@johnston_two_2020]. This study did not distinguish between cohabitation and marriage and revealed some further patterns in family structure changes, and where each pattern was intuitively interpretable. In particular, the latent class analysis distinguished classes on the basis of timing, where early and late entry into both a family with a sole mother as well as into a stepfamily became distinct types. It was also found that the different types identified both by sequence analysis and latent class analysis predicted the risk of later adolescent alcohol consumption. 

In order to understand why it might be of importance to approach children's family complexity through longitudinal profiles, I will devote the next section to life course theory. This framework will allow us to tie together what has so far been discussed: that (1) child outcomes depend both on experienced events and longer lasting states, (2) the childhood perspective provides something important beyond what we see when studying adults and (3) the need for taking into account the order and timing of events and states for understanding the potential consequences of family structural changes.

## Life course theory 

According to Elder et al. [-@elder_emergence_2003], life course theory came about from a "desire to understand social pathways, their developmental effect, and their relation to personal and social-historical conditions". Since its incipience in the 1970's, the *life course* has grown from a scientific idea to a theoretical framework and even a research field in its own right. With the launch of *Advances in Life Course Research* as a quarterly journal, it was declared that the life course is coming of age [@billari_life_2009]. 

The main theoretical concepts from the mature life course research which will be used in this thesis is those of *transitions*, *trajectories*,  *pathways* and *linked lives*. *Transitions* are state changes that are discrete and time-bound [@george_sociological_1993]. These can occur in multiple domains --- such as the social, biological, or economic --- and examples that have already been discussed are divorce and re-partnering. In turn, multiple transitions make up life *trajectories*. These consists of longer periods of change and stability, and although marked by a series of transitions they have a continuity of direction [@hutchison_life_2005]. A family formation trajectory by Hollywood blueprint would consist of dating followed by cohabitation and finally a wedding, marking a transition into married life. 

Just as trajectories are made up of a set of transitions, *pathways* are constituted by interlocking transitions and trajectories across time and domains [@macmillan_families_2005]. In particular, pathways are often in relation to entering and exiting social institutions such as a career or family. Although the combinatorial possibilities of transitions and trajectories are vast, the number of actualized pathways are usually limited and in relation to normative systems [@white_application_2014]. A way of thinking about pathways is that they are lived in relation to schema; that is, blueprints of culturally appropriate behavior, which are enacted more or less consciously [@sewell_theory_1992]. Schema inform individual heuristics for the temporal organization of the life course by providing a set of rules and models for how to link pathways over time [@macmillan_families_2005]. A modern Scandinavian pathway in relation to family formation would be forming a cohabiting union followed by having children, skipping the previously normative transition of marriage [@hofacker_patterns_2014]. 

Although often emphasized in the literature, the power of normative schema in shaping the individual life course has been contested [@macmillan_structure_2005]. Apart from norms, institutional and structural processes also shape lives and are able to produce homogeneity in life patterns [@marini_age_1984], as we have also seen in the discussion on the proliferation of the SDT. A further way of thinking about pathways, uncommitted to theoretical assumptions about their generation, is simply as a set of probabilities of transitioning between different states. But rather than a set of independent probabilities, they are path dependent with a certain degree of historicity, which can give rise to common patterns in a population [@rutter_pathways_1989]. 

Finally, a central idea in the life course framework is that of *linked lives*. This concept highlights the interdependence of our lives and that transitions that occur in one persons life often entail transitions for other people [@elder_emergence_2003]. This principle has particularly been applied in studies on the family and intergenerational relations [@carvalho_family_2021]. Already in one of the first studies using a life course perspective, @elder_children_1974 found that children's development could be impeded by the economic hardships experienced by their parents. This interdependence between parents and children, and how their social lives are intertwined, has since been explored in much depth and breadth. 

It is with this interdependence in mind, that I approach my object of study. Children's early family experiences are highly dependent on their parents' partnership trajectories, both in terms of vertical and horizontal family relations. At the same time, children's early family trajectories may themselves be associated with future outcomes. Thus, it is of interest to know how children's family trajectories are constituted. Before turning to how to conceptualize and operationalize these trajectories, I will turn to a conceptual model that can be used to understand why childhood family life course trajectories may differ between European countries. 

## Explaining cross European differences

One of the most adopted theories in explaining the macro level changes in family patterns during the late 20^th^ century is the Second Demographic Transition (SDT) theory. The SDT theory starts by identifying a set of phenomena, such as rising levels of cohabitation, living apart together arrangements, higher divorce rates, fertility postponement and an increase in out-of-wedlock childbirths [@lesthaeghe_second_2014]. In the early SDT theory, the changing family dynamics were seen as a result of ideational changes in Western Europe. In particular, the rise of postmodern attitudes and norms [@van_de_kaa_postmodern_2001], tightly linked to Inglehart's concept of a post-materialist society [@inglehart_silent_1971]. In particular, ideas such as individualism and self-realization are seen to have lead to both more individualized decisions on family formation, uncoupled from previous family norms, as well as a greater acceptance for issues such as homosexuality, abortion and non-marital childbearing. 

Ideational changes do not however, exist in a vacuum. In a recognition that also other social forces shape family decisions, Lestaeghe and Vanderhoeft proposed the conceptual model of _Ready, Willing, and Able_ [@lesthaeghe_ready_2001], adopted directly from @coale_demographic_1973. _Readiness_ refers to that new forms of behavior must come with some utility to the actor and _willingness_ to the acceptability of new patterns of action in relation to normative schemes, which is the aspect usually associated with the SDT. _Ability_ on the other hand, refers to if condition are in place to realize these new forms of behavior. In the case of family life, ability is related to, for instance, access to contraception and legal regulations of family relationships. This conceptual model bridges the relation between ideational changes, established normative systems as well as technological and legal systems. As both norms and institutional factors differ between countries, we would expect that even if changes in individual attitudes uniformly move towards more post-materialist ideals, the degree to which these attitudes can be realized will differ between countries and the processes have, as we shall see, been very different. 

Scandinavia has been a forerunner with regards to changing family forms. In Sweden, childbearing began to precede marriage already in the in the mid 1970's, when childbearing within cohabiting unions started to get mainstream acceptance [@thomson_partnership_2005]. At the same time, there was a drastic decrease in the level of total first marriage rates, dropping from about 0.9 in 1965 to 0.6 in 1975 [@sobotka_changing_2008]. In 2008, 60% of all children where born in non-marital cohabiting relationships [@olah_sweden_2008]. Drivers have been hypothesized to include both individualism and residential autonomy as well as institutional support also for childbearing outside of marriage [@olah_sweden_2008].

In other parts of Western Europe the patterns have been similar. The decrease in total first marriage rate started later than in Scandinavia, but fell to an even lower level in 2005 [@sobotka_changing_2008]. In the Netherlands in 2000, 55% of the Dutch population would opt for cohabitation before marriage as their ideal relationship trajectory [@fokkema_netherlands_2008], indicating an ideational shift towards cohabitation as preferable over direct marriage. Similarly, marriage has been decreasing in France and in 2005, the number of births in non-marital cohabitation where almost the same as the number of births in marital unions [@toulemon_france_2008]. Similar patterns are also found in Austria where 2% of women born 1941-45 had their first child in a cohabiting union, compared with 18% for women born 1966-67 [@prskawetz_austria_2008]. Further, over 25% of Austrian women born 1971-81 consider marriage to be an outdated institution [@prskawetz_austria_2008].

In Southern Europe, the drop in marriage rates started in the late 1970s and has not been as pronounced as in Northern and Western Europe [@sobotka_changing_2008]. Further, non-marital cohabitation is not nearly as common as in Northern and Western Europe [@de_rose_italy_2008;@delgado_spain_2008]. Does this mean that the cultural shift in values and norms did not occur in these countries? According to previous findings this does not seem to be the case for either Italy [@de_rose_italy_2008] or Spain [@delgado_spain_2008]. Rather, it seems that the economic situation of youth, coupled with a lack of active family policies is inhibiting changes in family dynamics. Thus, the young opt for staying in the parental home for longer as the option of cohabitation with a partner is not economically feasible, which in turn has led to a postponement in union formation [@baizan_political_2002]. The Southern European case seems to be a good example of where individuals display a high level of _willingness_, but lack both the resources (_readiness_) and proper support (_ability_) to change their family behaviors; which in turn gives rise to a different demographic pattern than in Northern and Western Europe.

Another region where there are clear mismatches between readiness, willingness and ability is in the Post-socialist countries. We first see evidence of the SDT in this region after the fall of the Iron Curtain in the early 1990s, when marriage rates started to drop steeply [@sobotka_changing_2008]. There is however a different timing when examining divorce, as already in the 1950's and 1960's parts of the Soviet block registered the highest divorce rates in Europe [@sobotka_changing_2008], indicating that divorce was an acceptable phenomena already in the early phase of the SDT in parts of Eastern Europe. It has also been found that tolerance towards pre-marital sex and divorce as well as values such as self-fulfillment outside of the family became salient in countries such as Poland [@kotowska_poland_2008] and Lithuania [@stankuniene_lithuania_2008] in the early 2000s. 

Tracing the developments in this region through the Readiness-Willingness-Ability approach, @sobotka_overview_2008-1 finds mixed indications in all three dimensions. The development of a social security net and the widespread employment of women increased readiness already before the fall of the Iron Curtain, but at the same time these jobs were often low-payed and females were still expected to take care of the household. As for willingness, the anti-religiousness of the Communist ideology helped break down traditional family norms and induced an atomization of society through the breakdown of religious and civic organizations [@speder_hungary_2008]. However, these new values were often adopted selectively [@sobotka_overview_2008-1]. Legal systems often enabled some of the family changes of the SDT, but contraceptive technology was often lacking showing that the ability to adopt new family behaviors was only possible to certain degrees.

Summing up the different regional patterns described above, two main pathways of the SDT have been identified and described [@sobotka_overview_2008-1]. The first path was driven by the classic SDT narrative of individualization and self-fulfillment, and primarily occurred among the affluent and highly educated. The second path arose as a response to changed structural conditions and economic crises, and primarily occurred among the less advantaged groups. Here there is also a regional divide, where the first pathway was primarily taken by countries in Northern, Western and to some degree Southern Europe, whilst Post-socialist countries mainly took the second. 

## Operationalizing family complexity over the life course

As has been outlined, the family structure of children may exhibit many different patterns and if we want to fully understand this complexity also within a life course framework, we need to study how these patterns play out over time. One fruitful analytic method which has been used within the life course literature is sequence analysis [for an overview see @liao_sequence_2022]. The foundation of sequence analysis lays in describing temporal processes as a series of states and state changes over discrete time units (such as months or years). It thereby provides a format for describing individual life course trajectories.

A usual approach for analyzing this type of data is through the use of cluster analysis [@piccarreta_holistic_2019]. This is a data-centric approach which deviates from traditional statistical methodologies in two main ways: 1) the units of analyses and 2) data assumptions. @aisenbrey_new_2010 show these differences by contrasting event history analysis with sequence analysis. Where event history analysis studies probabilities of transitions, sequence analysis studies patterns in trajectories as their main units of analysis. When it comes to assumption about the data, event history analysis assumes that the data is generated by a stochastic process whereas the algorithmic clustering makes no assumptions about the underlying process. This also leads to differences in practice, where event history analysis usually employs models where included covariates are derived from theory or previous empirical research, sequence analysis relies on algorithmic clustering of the data to identify groups of interest. These clusters are then used in two main ways: most commonly they are used as a dependent categorical variable in multinomial regression to find how other covariates predict life trajectories [@piccarreta_holistic_2019] but the clusters can also be used as predictor variables to see how life trajectories influence later life outcomes [e.g. @johnston_two_2020].

However, a fair amount of critique has been leveled against the clustering techniques employed by sequence analysis. To put this critique in methodological context, I will first outline the general steps in producing sequence analytic clusters before discussing the methodological drawbacks. The first step is to create a distance (or dissimilarity) matrix by calculating the distance between each pairwise combination of individual sequences in the data. Since the seminal work of @abbott_optimal_1986, the Optimal Matching algorithm has provided a standard way of calculating such a distance metric. The basic intuition is that we can permute a sequence by two kinds of operations to arrive at another sequence. The operations are substitutions (replacing one state by another) and insertions/deletions of states (commonly known as _indels_). In most cases, it is however not as straightforward as just summing up the number of operations needed to permute one sequence into another to get their distance. As the probability of transitions between states differ between pairs of states, or between substitution and indel operations, we need a substitution matrix which gives a cost to each possible substitution between pairs of states. We also need to assign a cost to indel operations. Thus, the optimal matching algorithm tries to find a set of operations that optimize the cost of permuting one sequence to another. This optimized cost is used as a measure of distance. Using the finalized distance matrix, it is possible to apply a clustering algorithm to derive a set of clusters, this is commonly done through Ward's method [@aisenbrey_new_2010]. This is an agglomerative hierarchical clustering algorithm that produces cluster by minimizing the within-cluster variance, using the squared dissimilarity distances as inputs [@murtagh_wards_2014].

The criticism leveled towards this methods is focused on two main areas: firstly, on the perceived lack of substantial results produced by the methodology and secondly, and perhaps more importantly, on the relationship between methodological assumptions and social processes. Both @wu_comments_2000 and @levine_but_2000 criticize sequence analysis for not delivering on early promises and that substantial results and new areas of research have not been borne out of it. As summarized in @liao_sequence_2022, there is a large variety of sociological articles being published using sequence analysis which would render that point of contention moot, but it does emphasis the importance of scrutinizing the sequence analysis methodology and if it is a valid tool for modeling or describing social processes.

@wu_comments_2000 offers a critique of sequence analysis where the most pressing is if a discrete sequence with events coupled with substitution and indel costs is a good model of life trajectories. The main problem lies in determining costs, as each cost is an assumption the researcher has to make, and that costs in the optimal matching algorithm are symmetric. This symmetric assumption means that transition from state $s_j \rightarrow s_k$ carries the same cost as transition from $s_k \rightarrow s_j$. But the question posed by Wu, is if the transitions between social states, such as employment and unemployment, are symmetrical in their probabilities or not. Which would imply that sequence representation through optimal matching is not isomorphic to the social processes they purport to represent. Further, Wu argues that as determining substitution and indel costs is up to the discretion of the researcher, it introduces more assumptions and potential biases in the analysis than proponents of sequence analysis had up to then acknowledged. This is a valid criticism but loses some of its relevance if we consider sequence analysis as an exploratory or descriptive analysis method, rather than a substitute to statistical modeling.

<!-- __TO DISCUSS__
Problems with clustering: @warren_different_2015 @han_comparing_2017 @von_luxburg_clustering_2010 -->

## Aims and expectations

The aim of this thesis is to further our understanding of children's family lives in Europe by examining them through the entire childhood life course. By using the profile based methods of sequence and cluster anlyses I aim to distill the diversity of family trajectories to a set of distinct family trajectory types. By doing this I will be able to compare the children of different countries and create a descriptive picture of childhood family lives in Europe and to further understand their similarities and differences. Given that within country differences are also to be expected, I will aim to see to what extend the mother's socioeconomic status is related to the family trajectory of her children. In particular I will try to answer the research questions:

  1. What are the most common family trajectories among European children, and how do they vary across countries? 
  2. How does the risk of experiencing a particular family trajectory differ by the mother's socioeconomic status, as measured by level of education?

Given the descriptive character of this paper I will not do any formal hypothesis testing but given the theories and previous research outlined above, I have some expectations of what will be seen in the data. As for the first research question, I expect that we will see the smallest variation in Southern Europe and the southern countries in post-Socialist Europe where a large majority of children will live their entire childhood in intact families. In the norhtern Post-Socialist countries, I expect there the be a fare share of non-intact family trajectories which will consist of both children born to lone mothers and to children of divorce. In Scandinavia I expect that trajectories driven by lone mothers will be rare, but that there will be a sizable share of children of divorce. Similar patterns are expected for Western Europe, perhaps with a smaller degree of stepfamilies and somewhat larger degree of single-mother families compared to Scandinavia. As for the second research question, I expect that the proportion of children with intact family histories will increase with increasing education. In particular, I expect that the incidence of non-intact families driven by birth to single mothers will decrease with education. 

# Data and Methods

## Analytical strategy

The first step in the analysis consists of defining family states and generating individual family sequences. As the data does not track both original parents of a child, I will only follow the mother's partnership trajectory with the assumption that children are residing at least part time with their mother during their childhood. Thus, joint physical custody is not taken into account, and the family trajectories on the father's side are not measured. This will make the measures of family complexity biased downwards. A positive note is that it will avoid the problem that men tend to under report their number of children in retrospective surveys [@juby_where_1999]. So although an analysis solely on the mother's side is a limitation, we can proceed as long as we keep that limitation in mind. Having constructed the parental structure sequence, I compute sequence dissimilarities and generate clusters for the full study population. The final number of clusters are selected using both cluster quality criteria and the substantive significance of the cluster division.

The problems of the validity of clustering typologies have been recognized for a long time, and methods for ameliorating this deficit have been developed within the bioinformatics literature [for an overview see @von_luxburg_clustering_2010], few sequence analysis studies within the social sciences have taken this issue to heart. In order to improve the validity of the typology, I will use a parametric bootstrap technique proposed by @hennig_flexible_2015 and implemented for sequence data by @studer_validating_2021 to aid with the selection of the final clusters.

Once the family typology is in place, I will show the proportion of the different family types for each country and region in the study, thus providing a descriptive of how children's family trajectories vary across Europe. This will also be broken down further by the mother's educational level.

## Data

The data used in this study comes from the birth and union histories in the Harmonized Histories data set [@perelli-harris_harmonized_2010]. This data set is compiled from a set of national surveys, primarily the Generations and Gender Survey, but also with other nationally representative surveys on partnerships and childbearing. The data is based on individual retrospective reports on partnership and childbearing histories.

The surveys were conducted differently in different countries. Generally, they were conducted through interviews [@fokkema_generations_2016], but self-filled web questionnaires were also used [@instituto_nacional_de_estadistica_fertility_2019]. Different sampling procedures were also used and more detailed information can be found in @fokkema_generations_2016 and @perelli-harris_harmonized_2010. Response rates are avaialble for most surveys and are presented below. 

In @tbl-1 I show basic information about the study population. For all countries in the study children born between 1980 and 1988 are included. For some countries such as Belgium, Netherlands, Poland, Spain and Sweden there are also children born well into the 1990's and the early 2000's. Each country, except Austria, is well represented with over 1,000 children. Belarus, Norway, Poland and Spain all have over 3,300 children included. That over 12,000 of the `r tar_read(final_data)[,.N]` respondents come from the Catholic countries Poland and Spain, might bias the analysis towards favouring clustering solutions which works well for these particular family contexts. Which is a discussion I will return to later.

```{r table1}
#| echo: false
#| results: "asis"
#| label: tbl-1
#| tbl-cap: "Descriptive information on survey years, response rates, cohorts included and number of children from each country. Some response rates are missing due to the sampling procedure of the surveys."

tar_read(table_1)
```

Other than that there is a well balanced sample, with good representation from all European regions, and enough respondents to perform subgroup analyses. 

I start of by excluding several countries from the data set. Germany is excluded due to its poor family history data and the UK is excluded due to oversampling respondents with low socioeconomic status. Uruguay, Kazakhstan, Canada and the U.S are removed as they are not European countries.

I then have `r tar_read(parents_tagged)[,.N]` total observations from `r length(tar_read(parents_tagged)[,unique(COUNTRY)])` different countries. I remove all men from the sample. Next, I remove childless women and respondents with missing birth information. I remove those where there is missing information on time of union or dissolution, overlapping relationships and those where a union dissolution is reported to have occurred before the union start. 

In order to be able to analyze the trajectories from the children's lived experience I reshape the data so that each child becomes a unit of observation. At the start I have `r tar_read(children_tagged)[,.N]` children. I remove children with missing birth month or where the birth is reported to have occurred before the mother is born. As I want to track children's' experiences until the end of age 15, I remove those in the sample who have not turned 16 at the time of interview as well as those who left home or died before their 16^th^ birthday. As I am interested in family changes after the advent of the Second Demographic Transition, I only keep those children born after 1980. In total `r tar_read(final_data)[,.N]` children, to `r tar_read(final_data)[unique(RESPID),.N]` mothers, remained in the sample. For a visual overview of the data funnel at each step, see Figure __x__ in the Appendix. 

## Sequence generation

For each child a family history was constructed on the maternal side, which is composed of the mother's partner history during the child's first 15 years of life. In order to construct this history, a STate-Sequence format [STS, @gabadinho_analyzing_2011] is used. This means that for each month of the child's life a column is generated which is filled with values describing the mother's union status and the child's number of siblings. The process time for analysis is from birth until turning 16, with quarterly precision. In STS format this is represented by 60 consecutive states, each capturing a three month period. 

Firstly, union spells were constructed. As there is no information on who is the father of the child if a mother was partnered at time of birth, that partner was assumed to be the other main parent. Then for each quarter, the representative state was coded as either *Single*, *Original parents* or *Step* depending on the mother's partner status in the first month of each quarter. The state is coded as *Single* if the mother is not in a union during the first month of the quarter. It is coded as *Original parents* if the mother is still in a relationship with the same partner as at the time of birth. Finally, a state is coded as *Step* if the mother is with a different partner than at time of birth. All consecutive partners are assumed to be new, i.e. no re-partnering with a former partner is assumed to occur. I do not distinguish between opposite-sex and same-sex partnerships, and the original parents are not necessarily the same as the biological parents, but simply constitute the partnership that the child is born into. Formally, the maternal union sequence can be described as the vector $\textbf{S}$ where  

$$
\textbf{S} = (s_1,s_2,s_3, \cdots, s_{60}) | s_i \in \{\textrm{Original parents, Single, Step} \}
$${#eq-sequence-example}

An example sequence of a child's first two years of life, where the mother is first partnered for 12 months (four quarters), then single for 6 (two quarters) and finally repartnered for 6 months (two quarters) is visible in @tbl-state-ex.

```{r}
#| echo: false
#| results: "asis"
#| label: tbl-state-ex
#| tbl-cap: "Two year union state sequence example"

months <- paste0("", 1:8)
states <- c(
  "OP", "OP", "OP", "OP",
  "Si", "Si", "St", "St"
)
knitr::kable(
  rbind(
    months,
    states
  ),
  "simple",
  row.names = FALSE
)
```

## Analysis

Having constructed the maternal union sequence, I calculate a dissimilarity matrix using an Dynamic Hamming Distance algorithm. This algorithm requires that the sequences are of the same lengths and does not use insertion and delation operation. A benefit of this approach is that it increases the sensitivity to sequence timing [@lesnard_setting_2010], which we know is important from the life course literature. It also means that indel costs needn't be defined which reduces the amount of decisions, and potentially biases, when generating the dissimilarity matrix. The substitution costs are derived from the transition rates in the data and thus reflects transition probabilities in this particular sample. This is the most data-centric approach possible for defining substitution costs and also reduces potential researcher biases, with the drawback of potentially reducing the external validity of the final clustering solution. 

This dissimilarity matrix is used to cluster the sequences, which is done using Ward's method. Being an agglomerative method, it starts with $k$ number of clusters where each observation in the sample is its own cluster, thus $k = N$ in the first step. The algorithm then progressively merges clusters until there is only one cluster containing all $N$ observations. Larger clusters are formed in steps by merging the two clusters with the shortest distance, where distance is defined by the increase in sum-of-squares if the clusters are merged. Given that this clustering method produces solution between $1$ and $N$ clusters, a choice regarding the final number of clusters needs to be made. This choice will be based on sociological relevance and cluster quality indices.

In choosing the number of clusters, it is of interest to maximize internal cluster homogeneity as well as between-cluster differences, and a common measure of this propriety is the Average Silhouette Width , introduced by @kaufman_finding_1990. This measure produces a value between $-1$ and $1$, where a larger value indicates a stronger structure. I also use Hubert's C , which captures the discrepancy between the partition obtained and the theoretically best partition, given the particular number of groups and their distances [@hubert_general_1976]. Hubert's C is defined in the range $[0,1]$ and a smaller value indicates a better partitioning. The issue with assessing cluster quality is that, unlike in traditional statistical modelling, there is no null model to compare the results with. So, even though @kaufman_finding_1990 propose an interpretation of the Average Silhouette Width measure, it cannot be generalized to all cluster applications. 

A potential solution is to use parametric bootstrapping to generate _null-model like_ distributions, cluster them, calculate a cluster quality index and compare it to the quality index value of the clusters from the actual data [@hennig_flexible_2015]. This has recently been applied to sequence analysis [@studer_validating_2021] and to my knowledge, this study will be one of the first to apply this technique. In practice, I generate a data set of 50 sequence sets that are similar but non-clustered using a combination of both sequencing and timing parameters [for the detail of the process see @studer_validating_2021]. Then the Average Silhouette Width and Hubert's C are calculated for each cluster solution ranging from 2 to 10 clusters for each of the 50 bootstraps. This produces a confidence interval for each of the quality measures for each $k$ grouping. This provides a baseline measure of the quality indices for data that is not clustered. If the index values of the cluster solutions for the actual data falls out of this bootstrapped range, it is a strong indicator that the particular clustering solution represents a "true" cluster and not just an artifact of the clustering process. However, that a cluster is true from a purely data-centric view, does not mean that it holds any form of social significance. Therefore, apart from using cluster quality indices, I will also look at the theoretical basis of the best performing cluster solutions when selecting the number of clusters to use. Further, I only consider 2 to 10 clusters as a larger number of cluster would be both unwieldy in terms of discussing the number of groups and the number of members in the smaller clusters would be too few to carry out any form of description on the country level. 

I provide a visual summary of the sequence clusters by a set of three plots for each cluster. First, a state distribution plot which shows the relative frequency of each state in the population by month. Second, a modal plot which shows the relative frequency of the most common state for each time point. Third, a medoid plot which shows the sequence with the smallest average dissimilarity to all other sequences in the group. It can be seen as a "median" sequence, or as the most typical trajectory within the cluster group. I also label the clusters with qualitative descriptions that aim to capture the general family experience of each cluster. 

In the next step, I provide a description of the proportions of children which belong to each cluster, on a country level. This allows us to detect similarities and differences between European countries as it relates to children's family experiences. In the final step I also relate parental union histories to the mother's educational level to see if there are also patterns stratified by socioeconomic status and if these are similar or different between countries.

The analysis is performed in R and relies on the packages **TraMineR** and  **WeightedCluster** [@studer_weightedcluster_2021] for the sequence and cluster analysis. 

## Ethical considerations

The Harmonized Histories data set is de-identified to protect the personal integrity of the study participants and all analysis presented are as aggregated results. In order to get access to the data I have signed a pledge of confidentiality, pledging not the share the data and only perform analysis related to the approved project. The data collection has been approved in each individual country and has been done under the oversight of the GGP Ethics Board and its Data Protection Office. I do not deal with any directly sensitive personal data such as ethnicity, political opinions or union membership. 

In order the improve both the transparancy and reproducibility of my research all code for generating the results, as well as the thesis, is available as a replication package at GitHub [@gustafsson_addegezonfamily_complexity_in_childhood_2022]. Anyone wishing the reproduce this research will have to apply for the Harmononized History dataset and then it will be possible to re-run my exact analyses using the replication package.  Using version control all changes in the project have been tracked so it is possible to follow the research process in detail. 


# Results

## Cluster diagnostics

The results from the cluster diagnostic bootstrap is presented in @fig-cwi. The pink shaded area represents the $95%$ confidence intervals of the standardized Average Sillhoute Width and Hubert's C and the purple line indicates the cluster quality of the 2 to 10 cluster solutions using the actual data. The graph shows that all cluster solutions in the range $[2,10]$ lay outside of the bootstrapped span, which in turn indicates that, from a data-centric view, there is always a significant clustering structure.

```{r}
#| echo: false
#| label: fig-cwi
#| fig-cap: "Standardized Average Silhouette Width and Hubert's C for actual and bootstrapped data"

tar_read(p_cluster_cqi)
```

We also see that the top scoring solutions for the Average Sillhoute Width are the 7- and 10-cluster solutions, and the 5- and 7-cluster solutions for Hubert's C. When examining the clusters visually as well as their size, I find that the 5-cluster solution is the most useful for this analysis. The 10-cluster solution is to large and some of the group sizes are too small for any analyses, and the same problem occurs for the 7-cluster solution. Working with five clusters still allows us to uncover heterogeneity in the complex family trajectories while still getting enough power to perform subgroup analyses (for visualisations of the 7- and 10-cluster solution consult the @fig-seven-cluster and @fig-ten-cluster in the Appendix).


## Final clusters

The 5-cluster solution is visualized in @fig-joint-plot. This is a compound figure which is composed of three different visualization for each cluster. The top plot is a sequence index plot, which displays all the individual sequences within a cluster as a horizontal line. For clarity, this plot is sorted by when the first state transition occurs. The middle plot, which in some cases looks quite similar, is a state distribution plot. Here the $y$-axis shows the proportion of cluster members who are in each state at time point $x$. It can be read as a series of stacked bar charts laid out over time. Finally, the bottom plot show the medoid sequence for each cluster. This is the sequence which has the shortest mean distance to all other sequences in the cluster. It can be interpreted as the most typical sequence in the set. The labels are my own qualitative descriptions of the clusters, and should be interpreted with caution and in relation to the information displayed.

```{r}
#| echo: false
#| label: fig-joint-plot
#| fig-cap: "Compound plots of the 5-cluster solution. For each cluster the topmost plot is a sequence index plot, the middle plot a state distribution plot and the bottom plot is a visualization of the medoid sequence. Each show different distributions of the states OP (Original Parents), Single (living with a single mother) and Step (living with mother and stepfather)."
#| fig-height: 8

tar_read(p_clusters_5) +
  theme(
    plot.title = element_text(size = 2),
  )
```

A first thing to note is that the "Intact original family" cluster is by far the largest one, with around `r round(tar_read(children_clusters)[cluster == "Intact original family",.N]/tar_read(children_clusters)[,.N] * 100, 0)`% of children in the full sample belonging to this cluster. On the opposite end of the spectrum we have the "Original family to stepfamily" cluster, which is composed of  `r round(tar_read(children_clusters)[cluster == "Original family to stepfamily",.N], 0)` children. Before showing the distribution of these clusters by country, I would like to spend some time to discuss each cluster and what information we can derive from these joint visualizations.

Starting of with the "Intact original family" cluster, we see that not all individuals actually live in an intact original family throughout childhood. A fair proportion (`r tar_read(children_clusters)[cluster == "Intact original family", round(sum(FAMILY_STATE177 != "OP")/.N, 2)]`) of children actually experience parental separation at some point, but all of them experience it after the age of 6. This highlights the need to be careful when infering group characteristics from sequence clusters.

The cluster "Original family to stepfamily" is composed of children who are all born into a parental union and spend at least some time in it before their parents divorce and eventually ending up in a stepfamily. In the index plot we can see that most children spend quite a short time with a single mother before she repartners, which is also highlighted in the medoid plot. In both the the index and state distribution plot, we see that some children end up living with a single mother again towards the end of the observed period. 

In the third plot, "Early separation", we observe a group of children where all are born into a parental union and the bulk experience a parental separation early on in childhood. A few experience stepfamily formation but most remain with a single mother throughout childhood. Here we can note that there is no "late separation" cluster, these have instead been classified with the "Intact original family" group.

The cluster "Single mother" consists mostly of children who spend their entire childhood with a single mother, but there are som who also live in a stepfamily or with their original parents in early childhood for a brief period of time. The fifth category, "Single mother to stepfamily", seems to contain some heterogeneity within the cluster where we see two common forms of family trajectories. The dominant one, which I've named the cluster after, consists of children being born to a single mother and who then enter a stepfamily before the age of 10. There is also a small group who are born into a union but experience a divorce shortly afterwards. 

In @fig-complex-proportion we see the proportion of children that are in each cluster, on a country by country basis, grouped by region. The left part of the figure shows the proportion of all children and the right hand figure shows the distrubtion over only those children who are not classified as "Intact original family". This is so that we can see in better detail how the distribution looks of these family sequence types. Countries are sorted within their respective region by the proportion of children beloning to the "Intact original family" group. 

```{r}
#| echo: false
#| label: fig-complex-proportion
#| fig-cap: "Proportion of children cluster by country, grouped by region. The right plot shows "
#| fig-height: 8

tar_read(p_combined_proportions)

```

```{r}
#| echo: false
#| label: fig-map-cluster
#| fig-cap: "Proportion of children cluster by country, grouped by region. The right plot shows "
#| fig-height: 10

tar_read(p_map_cluster) + theme(text = element_text(size = 8))

```

```{r}
#| echo: false
#| label: fig-map-edu
#| fig-cap: "Proportion of children cluster by country, grouped by region. The right plot shows "
#| fig-height: 10

tar_read(p_map_edu) + theme(text = element_text(size = 8))

```

# Discussion

## Limitations

# Conclusion

# Acknowledgments

# References

::: {#refs}
:::

# Appendix

```{r}
#| echo: false
#| results: "asis"
#| label: tbl-proportion-family-type
#| tbl-cap: "The proportion of children classified according to each family type by European country, with 95% confidence intervals in square brackets. The NA values arise because there are no Georgian children in the Mutiple single spells-cluster."

tar_read(tab_proportion_family_type) |> 
  kableExtra::kable_styling(font_size = 7) |>
  kableExtra::column_spec(2:8, width = "1.7cm") 
```

```{r}
#| echo: false
#| label: fig-seven-cluster
#| fig-cap: "5-cluster solution"
#| fig-height: 8

tar_read(p_clusters_7) +
  theme(
    plot.title = element_text(size = 2),
  )
```

```{r}
#| echo: false
#| label: fig-ten-cluster
#| fig-cap: "10-cluster solution"
#| fig-height: 8

tar_read(p_clusters_10) +
  theme(
    plot.title = element_text(size = 2),
  )
```